---
title: "Architecture Overview"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, dplyr::rename, .quiet = TRUE)
```

## Class hierarchy

PKNCA uses four S3 classes that form a strict pipeline. Each wraps the previous one:

```{mermaid}
classDiagram
    class PKNCAconc {
        data: data.frame
        formula: conc~time|groups
        subject: column name
        sparse: logical
        units: concu/timeu
        exclude / exclude_half.life
    }
    class PKNCAdose {
        data: data.frame
        formula: dose~time|groups
        route: intravas./extravas.
        duration / rate
        units: doseu/timeu
    }
    class PKNCAdata {
        conc: PKNCAconc
        dose: PKNCAdose
        intervals: data.frame
        impute: method string
        units: pknca_units_table()
        options: named list
    }
    class PKNCAresults {
        result: data.frame
        data: PKNCAdata
    }

    PKNCAconc --> PKNCAdata
    PKNCAdose --> PKNCAdata
    PKNCAdata --> PKNCAresults
```

## The interval column system

The parameter registry is the core extensibility mechanism. Every NCA parameter is registered as an **interval column** via `add.interval.col()`.

```{r}
# Inspect registration for auclast
cols <- get.interval.cols()
str(cols[["auclast"]])
```

Each entry contains:

| Field | Purpose |
|---|---|
| `FUN` | The R function that computes this parameter |
| `depends` | Character vector of parameters that must be computed first |
| `desc` | Human-readable description |
| `formalsmap` | Maps function arguments to interval/group/options data |
| `value.ok` | Validation function for the computed result |
| `sparse` | Whether this parameter works for sparse PK |

## formalsmap

`formalsmap` is how PKNCA passes the right data to each parameter function. Arguments can be sourced from four places:

| Source type | Example | What it provides |
|---|---|---|
| Interval column | `conc`, `time` | Concentrations/times for the current interval |
| Other parameters | `auclast = "auclast"` | A previously computed NCA result |
| Interval bounds | `start`, `end` | The interval start/end times |
| Options | `auc.method` | Value from `PKNCA.options()` |

## Writing a custom parameter

To add a new NCA parameter:

```{r}
# Example: compute the "exposure ratio" — AUCinf / AUClast
pk.calc.exposure.ratio <- function(aucinf.obs, auclast) {
  aucinf.obs / auclast
}

add.interval.col(
  name     = "exposure.ratio",
  FUN      = "pk.calc.exposure.ratio",  # must be a character string, not the function object
  unit_type = "unitless",
  pretty_name = "Exposure Ratio (AUCinf/AUClast)",
  desc     = "Ratio of AUCinf to AUClast; >1 indicates extrapolation",
  depends  = c("aucinf.obs", "auclast"),
  formalsmap = list(aucinf.obs = "aucinf.obs", auclast = "auclast")
)

# Now it can be requested in intervals
d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose <- Theoph |> as.data.frame() |>
  dplyr::group_by(Subject) |>
  dplyr::summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  dplyr::rename(subject = Subject) |>
  dplyr::mutate(time = 0)

o_conc <- PKNCAconc(d_conc, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")

custom_interval <- data.frame(
  start = 0, end = Inf,
  auclast = TRUE, aucinf.obs = TRUE,
  exposure.ratio = TRUE
)

o_data <- PKNCAdata(o_conc, o_dose, intervals = custom_interval)
o_nca  <- pk.nca(o_data)

as.data.frame(o_nca) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "aucinf.obs", "exposure.ratio")) |>
  dplyr::select(subject, PPTESTCD, PPORRES) |>
  dplyr::arrange(subject, PPTESTCD)
```

## Options system

`PKNCA.options()` is a named list stored as a package-level environment. It is accessed in three ways:

```{r}
# 1. Read all options
PKNCA.options()

# 2. Read a specific option
PKNCA.options("min.hl.points")

# 3. Set globally (affects all subsequent analyses in the session)
old <- PKNCA.options(min.hl.points = 4)
PKNCA.options("min.hl.points")

# 4. Restore previous value — must use do.call, not PKNCA.options(old) directly
do.call(PKNCA.options, old)
PKNCA.options("min.hl.points")
```

> Prefer passing `options = list(...)` to `PKNCAdata()` over mutating global options. Global changes persist across analyses in the same session.

## Sparse PK

Sparse PK (one or two samples per subject, common in preclinical or paediatric studies) is handled via a parallel code path:

- Set `sparse = TRUE` in `PKNCAconc()`
- Concentrations are pooled across subjects per timepoint
- Mean concentration-time profile is computed, then NCA is run on that profile
- Supported parameters are flagged with `sparse = TRUE` in their `add.interval.col()` registration

## Key internal functions

| Function | Purpose |
|---|---|
| `get.interval.cols()` | Return full parameter registry |
| `PKNCA:::sort.interval.cols()` | Topological sort → execution order |
| `PKNCA:::pk.nca.interval()` | Compute all parameters for one interval/group |
| `pknca_units_table()` | Build unit assignment/conversion table |
| `PKNCA_impute_method_*()` | Built-in imputation methods |
| `pk.calc.auc.last()` | Core AUC calculation (trapezoidal) |
| `pk.calc.half.life()` | λz regression with best-fit selection |

---

> **Depth note (for Teun):** This overview covers the surface. The formalsmap system, business rules (`002-pk.business.rules.R`), and the validation framework (`v60-PKNCA-validation.Rmd`) are richer topics — revisit when needed.
