---
title: "NCA Parameter Dependencies"
---

```{r setup, include=FALSE}
library(PKNCA)
library(visNetwork)
library(dplyr)
```

## Overview

PKNCA resolves parameter dependencies automatically. When you request `aucinf.obs`, PKNCA internally ensures `lambda.z` is computed first because `aucinf.obs` depends on it. This graph makes those dependencies explicit.

## How dependencies are defined

Every NCA parameter is registered via `add.interval.col()` with a `depends` vector listing the parameters that must be computed before it. The full registry is accessible via `get.interval.cols()`.

## Interactive dependency graph

```{r build-graph}
# Get all registered interval columns (parameters)
cols <- get.interval.cols()

# Build edge list from the 'depends' field
edges <- lapply(names(cols), function(param) {
  deps <- cols[[param]]$depends
  if (length(deps) == 0) return(NULL)
  data.frame(from = deps, to = param, stringsAsFactors = FALSE)
}) |>
  bind_rows()

# All unique nodes
all_nodes <- unique(c(edges$from, edges$to, names(cols)))

# Assign category for colouring
categorise <- function(name) {
  case_when(
    grepl("^auc",       name) ~ "AUC",
    grepl("^lambda|half\\.life|r\\.squared", name) ~ "Half-life / λz",
    grepl("^cl\\.|^vz\\.|^vss", name) ~ "CL / Volume",
    grepl("^mrt",       name) ~ "MRT",
    grepl("^c(max|min|0|last|first)|^t(max|min|last|first|lag)|clast", name) ~ "Cmax / Tmax",
    grepl("^f$|bioavail|^ae|^fe|^clr|^vd",  name) ~ "Bioavailability / Renal",
    grepl("^tss",       name) ~ "Steady-state",
    TRUE ~ "Other"
  )
}

category_colours <- c(
  "AUC"                    = "#4e79a7",
  "Half-life / λz"         = "#f28e2b",
  "CL / Volume"            = "#e15759",
  "MRT"                    = "#76b7b2",
  "Cmax / Tmax"            = "#59a14f",
  "Bioavailability / Renal"= "#edc948",
  "Steady-state"           = "#b07aa1",
  "Other"                  = "#aaaaaa"
)

nodes <- data.frame(
  id    = all_nodes,
  label = all_nodes,
  group = categorise(all_nodes),
  stringsAsFactors = FALSE
) |>
  mutate(color = category_colours[group])

visNetwork(nodes, edges,
           width = "100%", height = "700px",
           main = "PKNCA NCA Parameter Dependencies") |>
  visNodes(shape = "dot", size = 14,
           font = list(size = 12),
           borderWidth = 2) |>
  visEdges(arrows = "to",
           color  = list(color = "#aaaaaa", highlight = "#333333"),
           smooth = list(type = "cubicBezier")) |>
  visOptions(
    highlightNearest = list(enabled = TRUE, degree = 2, hover = TRUE),
    nodesIdSelection = list(enabled = TRUE, main = "Select parameter")
  ) |>
  visLegend(useGroups = TRUE, position = "right") |>
  visPhysics(stabilization = list(iterations = 200)) |>
  visLayout(randomSeed = 42)
```

**How to use this graph:**

- Click a parameter node to highlight its immediate dependencies and dependents
- The "Select parameter" dropdown jumps to a specific node
- Drag nodes to rearrange; scroll to zoom
- Colours indicate parameter family (see legend)

## Dependency table

For a text-based reference:

```{r dep-table}
dep_table <- lapply(names(cols), function(param) {
  data.frame(
    parameter = param,
    depends_on = if (length(cols[[param]]$depends) == 0) "(none)"
                 else paste(cols[[param]]$depends, collapse = ", "),
    stringsAsFactors = FALSE
  )
}) |>
  bind_rows() |>
  arrange(parameter)

knitr::kable(dep_table, col.names = c("Parameter", "Depends on"))
```

## Dependency resolution

When you call `pk.nca()`, PKNCA calls `sort.interval.cols()` internally to produce a topologically sorted execution order — parameters with no dependencies run first, then those that depend on already-computed values.

```{r sort-order}
# Show the resolved execution order
sorted <- PKNCA:::sort.interval.cols(cols)
data.frame(
  order     = seq_along(sorted),
  parameter = names(sorted)
) |>
  head(30) |>
  knitr::kable(caption = "First 30 parameters in resolved execution order")
```
