---
title: "AUC Types and Integration Methods"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)

# Shared dataset for this page: oral theophylline, subject 1
d_one <- as.data.frame(Theoph) |>
  filter(Subject == "1") |>
  rename(time = Time, subject = Subject)

d_dose <- data.frame(subject = "1", dose = Theoph$Dose[1] * Theoph$Wt[1], time = 0)

o_conc <- PKNCAconc(d_one, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")
```

## AUC types

PKNCA offers four families of AUC, each answering a different question:

```{mermaid}
flowchart LR
    A["AUClast\n0 → last measurable"]
    B["AUCall\n0 → last, zeros for BLQ"]
    C["AUCinf.obs\n0 → ∞ (observed Clast)"]
    D["AUCinf.pred\n0 → ∞ (predicted Clast)"]
    E["AUCint / partial\n0 → t (any t)"]
```

| Parameter | End point | BLQ treatment | When to use |
|---|---|---|---|
| `auclast` | Last measurable concentration | BLQs after last non-BLQ are ignored | Default; most studies |
| `aucall` | Last measured time point | Post-last-measurable BLQs treated as 0 | When you want to include the BLQ tail |
| `aucinf.obs` | Extrapolated to ∞ (using observed Clast) | — | When half-life is reliable |
| `aucinf.pred` | Extrapolated to ∞ (using predicted Clast from λz fit) | — | When Clast is noisy |
| `aucint.*` | Any user-defined time window | — | Partial AUC, bioequivalence |

### AUClast vs AUCall

```{r}
auc_compare <- data.frame(
  start    = 0,
  end      = Inf,
  auclast  = TRUE,
  aucall   = TRUE,
  aucinf.obs = TRUE
)

o_data <- PKNCAdata(o_conc, o_dose, intervals = auc_compare)
o_nca  <- pk.nca(o_data)

as.data.frame(o_nca) |>
  filter(PPTESTCD %in% c("auclast", "aucall", "aucinf.obs", "aucpext.obs")) |>
  select(PPTESTCD, PPORRES)
```

> `aucall` ≥ `auclast` when there are post-last-measurable BLQ timepoints, since those are treated as 0 and add a small triangle to the AUC.

---

## Partial AUC (AUCint)

For bioequivalence or exposure-in-a-window calculations, use `aucint.*` parameters.
These calculate AUC over an exact time window, interpolating concentrations at the boundaries if needed.

```{r}
# AUC from 0 to 4h and 0 to 12h on the same profile
partial_intervals <- data.frame(
  start         = c(0,  0),
  end           = c(4, 12),
  aucint.last   = TRUE,   # AUC to the last obs within window (or interpolated boundary)
  aucint.inf.obs = FALSE
)

o_data_partial <- PKNCAdata(o_conc, o_dose, intervals = partial_intervals)
o_nca_partial  <- pk.nca(o_data_partial)

as.data.frame(o_nca_partial) |>
  filter(PPTESTCD == "aucint.last") |>
  select(start, end, PPTESTCD, PPORRES)
```

**All partial AUC variants:**

| Parameter | End concentration | Extrapolation |
|---|---|---|
| `aucint.last` | Last observed in window | None |
| `aucint.all` | Last observed, BLQs as 0 | None |
| `aucint.inf.obs` | Extrapolated to ∞ (observed Clast) | λz-based tail |
| `aucint.inf.pred` | Extrapolated to ∞ (predicted Clast) | λz-based tail |

```{r}
# All partial AUC variants over a single window
all_partial <- data.frame(
  start            = 0,
  end              = 8,
  aucint.last      = TRUE,
  aucint.all       = TRUE,
  aucint.inf.obs   = TRUE,
  aucint.inf.pred  = TRUE
)

o_nca_all_partial <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = all_partial))
as.data.frame(o_nca_all_partial) |>
  filter(grepl("^aucint", PPTESTCD)) |>
  select(PPTESTCD, PPORRES)
```

---

## Percent extrapolation (%AUCextrap)

When using `aucinf`, you should check what fraction was extrapolated beyond the last observation.
A high `aucpext` means the terminal phase was poorly characterised.

```{r}
pext_interval <- data.frame(
  start        = 0,
  end          = Inf,
  auclast      = TRUE,
  aucinf.obs   = TRUE,
  aucpext.obs  = TRUE,
  aucpext.pred = TRUE
)

o_nca_pext <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = pext_interval))
as.data.frame(o_nca_pext) |>
  filter(grepl("^auc", PPTESTCD)) |>
  select(PPTESTCD, PPORRES)
```

The default threshold is `max.aucinf.pext = 20` (%). Results where `aucpext > 20%` are flagged with an `exclude` message. Change the threshold via options:

```{r}
# Stricter: flag if more than 10% is extrapolated
o_data_strict <- PKNCAdata(o_conc, o_dose, intervals = pext_interval,
                            options = list(max.aucinf.pext = 10))
o_nca_strict  <- pk.nca(o_data_strict)
as.data.frame(o_nca_strict) |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(PPTESTCD, PPORRES, exclude)
```

---

## Integration methods

The integration method controls how the area of each trapezoid is computed between consecutive timepoints.

```{r}
# Visual comparison of the three methods on one subject
methods <- c("linear", "lin-log", "lin up/log down")

results <- lapply(methods, function(m) {
  o <- pk.nca(PKNCAdata(o_conc, o_dose,
                        intervals = data.frame(start=0, end=Inf, auclast=TRUE),
                        options   = list(auc.method = m)))
  as.data.frame(o) |>
    filter(PPTESTCD == "auclast") |>
    mutate(method = m)
}) |> bind_rows()

results |> select(method, PPORRES) |>
  rename(auclast = PPORRES)
```

### Method details

**Linear trapezoidal (`"linear"`)**

Uses the standard trapezoidal rule throughout:
$$\text{AUC}_{t_i \to t_{i+1}} = \frac{(C_i + C_{i+1})}{2} \times (t_{i+1} - t_i)$$

Overestimates during the descending phase (gives too much weight to the higher concentration).

**Lin-log (`"lin-log"`)**

Uses log-linear interpolation throughout the descending phase:
$$\text{AUC}_{t_i \to t_{i+1}} = \frac{(C_i - C_{i+1})}{\ln C_i - \ln C_{i+1}} \times (t_{i+1} - t_i)$$

Cannot handle zero or negative concentrations. Tends to underestimate the ascending phase.

**Lin up / log down (`"lin up/log down"`) — default**

Hybrid: linear trapezoidal on the ascending phase (when $C_{i+1} \geq C_i$), log-linear on the descending phase (when $C_{i+1} < C_i$). Best of both worlds for most PK profiles.

```{r}
# Visualise the three profiles for one subject
d_plot <- d_one |>
  filter(conc > 0) |>
  arrange(time)

# Illustrate the interpolated concentration between two points
t_seq <- seq(min(d_plot$time), max(d_plot$time), length.out = 200)

ggplot(d_plot, aes(x = time, y = conc)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(size = 3, colour = "steelblue") +
  geom_area(alpha = 0.15, fill = "steelblue") +
  labs(title = "Concentration-time profile (shaded area = AUClast)",
       x = "Time (h)", y = "Concentration (mg/L)") +
  theme_minimal()
```

### When to use which method

| Study type | Recommended method |
|---|---|
| Most oral/extravascular | `"lin up/log down"` (default) |
| IV infusion with smooth decline | `"lin up/log down"` |
| Regulatory (some agencies) | `"linear"` |
| Highly irregular profiles with zeros | `"linear"` |
| Never recommended | `"lin-log"` (numerical issues) |

---

## AUMC — area under the first moment curve

The AUMC is used to compute MRT. It weights concentrations by time:
$$\text{AUMC} = \int_0^{t} t \cdot C(t) \, dt$$

```{r}
aumc_interval <- data.frame(
  start        = 0,
  end          = Inf,
  aumclast     = TRUE,
  aumcinf.obs  = TRUE
)

o_nca_aumc <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = aumc_interval))
as.data.frame(o_nca_aumc) |>
  filter(grepl("^aumc", PPTESTCD)) |>
  select(PPTESTCD, PPORRES)
```

---

## Concentration interpolation at interval boundaries

When a requested interval boundary (e.g., `end = 8`) falls between two observed timepoints, PKNCA interpolates the concentration at that boundary before computing AUC.

The interpolation method matches the AUC method:
- `"linear"` → linear interpolation
- `"lin up/log down"` → log-linear interpolation on the descending phase

```{r}
# end=5h falls between t=4 and t=7.02 in the Theoph dataset
interp_interval <- data.frame(start = 0, end = 5, auclast = TRUE)
o_nca_interp <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = interp_interval))
as.data.frame(o_nca_interp) |>
  filter(PPTESTCD == "auclast") |>
  select(start, end, PPTESTCD, PPORRES)
```
