---
title: "Extravascular (Oral/SC) Examples"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## The EV dataset: Theophylline

We use `datasets::Theoph` — oral theophylline (bronchodilator) in 12 subjects, sampled over 25 hours after a single oral dose.

```{r}
head(Theoph)
str(Theoph)
```

Columns: `Subject`, `Wt` (body weight, kg), `Dose` (mg/kg), `Time` (h), `conc` (mg/L).

```{r}
# Prepare concentration data
d_conc <- as.data.frame(Theoph) |>
  rename(time = Time, subject = Subject, conc = conc)

# Prepare dose data: convert mg/kg to mg per subject
d_dose <- Theoph |>
  as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)

# Visualize concentration profiles
ggplot(d_conc, aes(x = time, y = conc, group = subject, colour = subject)) +
  geom_line() + geom_point() +
  labs(title = "Theophylline oral — concentration-time profiles",
       x = "Time (h)", y = "Concentration (mg/L)") +
  theme_minimal()
```

The typical absorption-distribution-elimination shape is visible: concentrations rise to a peak (Tmax/Cmax) then decline.

---

## Basic extravascular analysis

```{r}
o_conc <- PKNCAconc(d_conc, conc ~ time | subject)

o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")

# Core EV parameters
ev_intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE,
  lambda.z    = TRUE,
  cl.obs      = TRUE,   # apparent CL (= CL/F for EV)
  vz.obs      = TRUE    # apparent Vz (= Vz/F for EV)
)

o_data <- PKNCAdata(o_conc, o_dose, intervals = ev_intervals)
o_nca  <- pk.nca(o_data)

as.data.frame(o_nca) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

---

## Parameter reference

### Cmax and Tmax

Cmax is the maximum observed concentration. Tmax is the time at which it occurs.

```{r}
as.data.frame(o_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

**When there are multiple peaks** (fed state, enterohepatic recirculation), the default is to return the first Tmax. Control this with `first.tmax`:

```{r}
# Use the last Tmax when there are ties
o_data_lasttmax <- PKNCAdata(
  o_conc, o_dose,
  intervals = ev_intervals,
  options = list(first.tmax = FALSE)
)
```

### AUC variants

Same variants as IV, but note that for EV routes **CL and Vz are apparent** (divided by F, the bioavailability fraction):

| Parameter | Meaning |
|---|---|
| `auclast` | AUC from 0 to last measurable concentration |
| `aucinf.obs` | AUC extrapolated to ∞ (using observed Clast) |
| `aucinf.pred` | AUC extrapolated to ∞ (using predicted Clast) |
| `aucpext.obs` | % of AUCinf that is extrapolated |
| `cl.obs` | Apparent clearance = Dose / AUCinf.obs (this is CL/F) |
| `vz.obs` | Apparent volume = CL.obs / λz (this is Vz/F) |

```{r}
auc_params <- data.frame(
  start        = 0,
  end          = Inf,
  auclast      = TRUE,
  aucinf.obs   = TRUE,
  aucinf.pred  = TRUE,
  aucpext.obs  = TRUE,
  cl.obs       = TRUE
)

o_data_auc <- PKNCAdata(o_conc, o_dose, intervals = auc_params)
o_nca_auc  <- pk.nca(o_data_auc)

as.data.frame(o_nca_auc) |>
  filter(PPTESTCD %in% c("auclast", "aucinf.obs", "aucpext.obs", "cl.obs")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

### Half-life and λz

For EV, the terminal phase reflects elimination (not absorption) once absorption is complete. The same quality controls apply as for IV.

```{r}
hl_params <- data.frame(
  start             = 0,
  end               = Inf,
  lambda.z          = TRUE,
  half.life         = TRUE,
  lambda.z.n.points = TRUE,
  r.squared         = TRUE,
  adj.r.squared     = TRUE
)

o_data_hl <- PKNCAdata(o_conc, o_dose, intervals = hl_params)
o_nca_hl  <- pk.nca(o_data_hl)

as.data.frame(o_nca_hl) |>
  filter(PPTESTCD %in% c("half.life", "lambda.z.n.points", "adj.r.squared")) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

**Allow or exclude Tmax from the terminal regression:**

By default, the Tmax point is excluded from the λz regression (absorption may still be ongoing). You can override this:

```{r}
o_data_allow_tmax <- PKNCAdata(
  o_conc, o_dose,
  intervals = hl_params,
  options = list(allow.tmax.in.half.life = TRUE)
)
```

### Lag time (tlag)

`tlag` is the time before absorption begins — the delay between dosing and the first measurable rise in concentration. Common for enteric-coated formulations or SC injections with a diffusion delay.

```{r}
o_nca_tlag <- pk.nca(PKNCAdata(o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, tlag = TRUE, tmax = TRUE, cmax = TRUE)))

as.data.frame(o_nca_tlag) |>
  filter(PPTESTCD %in% c("tlag", "tmax", "cmax")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

> For the Theoph dataset tlag is 0 for all subjects (no lag). It becomes non-zero when the concentration-time profile shows a flat period after dosing before rising.

### Bioavailability (f)

`f` computes the ratio of AUC between two routes or treatments within the same subject. It requires a **reference** AUC in the grouping structure (typically set up with a treatment or period column).

```{r}
# Syntax: request f = TRUE; PKNCA computes AUC_test / AUC_reference
# Requires a grouping column that identifies test vs. reference
# Example structure (illustrative — Theoph is single-arm):
data.frame(start = 0, end = Inf, aucinf.obs = TRUE, f = FALSE)
# f = TRUE only makes sense when the data has paired test/reference periods
```

> See `?pk.calc.f` for the full setup. `f` is typically used in crossover bioequivalence studies.

### Volume at steady state (Vss)

Vss is computed from CL and MRT. For EV routes these are apparent values (divided by F).

```{r}
vss_params <- data.frame(
  start    = 0,
  end      = Inf,
  vss.obs  = TRUE,   # apparent Vss = CL.obs × MRT.obs
  vss.pred = TRUE
)

o_nca_vss <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = vss_params))
as.data.frame(o_nca_vss) |>
  filter(PPTESTCD %in% c("vss.obs", "vss.pred")) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

### Effective half-life and kel

```{r}
eff_params <- data.frame(
  start          = 0,
  end            = Inf,
  thalf.eff.obs  = TRUE,   # ln(2) × MRT — accumulation half-life
  thalf.eff.pred = TRUE,
  kel.obs        = TRUE    # elimination rate constant (= λz)
)

o_nca_eff <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = eff_params))
as.data.frame(o_nca_eff) |>
  filter(PPTESTCD %in% c("thalf.eff.obs", "thalf.eff.pred", "kel.obs")) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

### AUC for a specific tau (dosing interval)

For multiple-dose studies, request AUC over the dosing interval (tau):

```{r}
# Single dose, but demonstrating syntax for AUC(0-tau)
tau_interval <- data.frame(
  start      = 0,
  end        = 24,      # tau = 24 h
  auclast    = TRUE,
  cmax       = TRUE,
  tmax       = TRUE
)

o_data_tau <- PKNCAdata(o_conc, o_dose, intervals = tau_interval)
o_nca_tau  <- pk.nca(o_data_tau)
as.data.frame(o_nca_tau) |> select(subject, PPTESTCD, PPORRES)
```

---

## Imputation of missing predose concentrations

In some studies, the predose sample is missing or was not collected. PKNCA can impute a time-0 concentration before calculations.

Available built-in methods (comma-separate to chain them):

| Method | What it does |
|---|---|
| `start_predose` | Adds concentration = 0 at time 0 if no predose sample exists |
| `start_conc0` | Sets the predose concentration to 0 unconditionally |
| `start_cmin` | Uses the minimum observed concentration as predose value |

```{r}
# Remove the time=0 observation from subject 1 to simulate missing predose
d_conc_missing <- d_conc |>
  filter(!(subject == "1" & time == 0))

o_conc_miss <- PKNCAconc(d_conc_missing, conc ~ time | subject)

# With imputation: add back a 0 at time 0
o_data_imputed <- PKNCAdata(
  o_conc_miss, o_dose,
  intervals = ev_intervals,
  impute = "start_predose"
)

o_nca_imputed <- pk.nca(o_data_imputed)

# Compare subject 1 AUC with and without imputation
bind_rows(
  as.data.frame(o_nca) |> filter(subject == "1", PPTESTCD == "auclast") |> mutate(version = "original"),
  as.data.frame(o_nca_imputed) |> filter(subject == "1", PPTESTCD == "auclast") |> mutate(version = "imputed")
) |>
  select(version, PPTESTCD, PPORRES)
```

**Per-interval imputation** — apply different methods to different calculation windows:

```{r}
# Intervals with an "impute" column specifying method per row
d_intervals_impute <- data.frame(
  start   = 0,
  end     = Inf,
  auclast = TRUE,
  impute  = "start_predose"
)

o_data_per_interval <- PKNCAdata(
  o_conc_miss, o_dose,
  intervals = d_intervals_impute,
  impute = "impute"   # tells PKNCA to look in the intervals column named "impute"
)
```

---

## Bioavailability

Bioavailability (F) compares AUC from the EV route to AUC from an IV reference.
PKNCA calculates this if you supply both an IV and an EV `PKNCAresults` object.

```{r}
# IV reference (using Indometh as stand-in — illustrative syntax only)
d_iv_conc <- as.data.frame(Indometh)
d_iv_dose <- data.frame(Subject = unique(d_iv_conc$Subject), dose = 25, time = 0)

o_iv_conc <- PKNCAconc(d_iv_conc, conc ~ time | Subject)
o_iv_dose <- PKNCAdose(d_iv_dose, dose ~ time | Subject, route = "intravascular")
o_iv_data <- PKNCAdata(o_iv_conc, o_iv_dose, intervals = data.frame(
  start = 0, end = Inf, aucinf.obs = TRUE
))
o_iv_nca <- pk.nca(o_iv_data)

# pk.business.rule: bioavailability is computed via pk.calc.f()
# Use pk.calc.f() directly when you have matched IV and EV AUCs
iv_aucinf <- as.data.frame(o_iv_nca) |>
  filter(PPTESTCD == "aucinf.obs") |>
  summarise(mean_aucinf = mean(PPORRES)) |>
  pull(mean_aucinf)

ev_aucinf <- as.data.frame(o_nca_auc) |>
  filter(PPTESTCD == "aucinf.obs") |>
  summarise(mean_aucinf = mean(PPORRES)) |>
  pull(mean_aucinf)

# F (%) = (AUC_ev / Dose_ev) / (AUC_iv / Dose_iv) * 100
# These datasets use different drugs, so this is illustrative only:
cat("Illustrative F calculation:\n")
cat(sprintf("  IV AUCinf (mean): %.2f\n", iv_aucinf))
cat(sprintf("  EV AUCinf (mean): %.2f\n", ev_aucinf))
```

> For a real bioavailability study, subjects would receive both IV and EV treatments, and `pk.calc.f()` handles the ratio calculation accounting for dose scaling.

---

## Multiple-dose / steady-state

For multiple-dose studies, specify the dosing interval in your intervals data frame. PKNCA will compute accumulation-relevant parameters:

```{r}
# Steady-state interval parameters
ss_params <- c("cmax", "cmin", "tmax", "auclast", "half.life", "lambda.z")

ss_intervals <- data.frame(
  start  = 0,
  end    = 24,
  cmax   = TRUE,
  cmin   = TRUE,
  tmax   = TRUE,
  auclast = TRUE
)

o_data_ss <- PKNCAdata(o_conc, o_dose, intervals = ss_intervals)
o_nca_ss  <- pk.nca(o_data_ss)

as.data.frame(o_nca_ss) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

---

## Excluding observations

Mark individual time points to exclude (e.g., emesis within 2× Tmax, suspected contamination):

```{r}
d_conc_excl <- d_conc |>
  mutate(
    exclude_reason = ifelse(subject == "5" & time == 7.02,
                            "vomiting within 2h of Tmax", NA_character_)
  )

o_conc_excl <- PKNCAconc(d_conc_excl, conc ~ time | subject, exclude = "exclude_reason")

o_data_excl <- PKNCAdata(o_conc_excl, o_dose, intervals = ev_intervals)
o_nca_excl  <- pk.nca(o_data_excl)
```

Excluded points appear flagged in `as.data.frame()` output, preserving audit trail.

---

## BLQ (below limit of quantification) handling

PKNCA's `conc.blq` option controls how BLQ values (typically entered as 0) are treated:

```{r}
# Default: keep BLQ as 0
# "drop": treat BLQ as missing (NA)
# "ignore": use 0 as-is
# List form for per-interval control:

o_data_blq <- PKNCAdata(
  o_conc, o_dose,
  intervals = ev_intervals,
  options = list(
    conc.blq = list(
      before.tmax = "keep",   # keep 0s before Tmax (ascending phase)
      after.tmax  = "drop"    # drop BLQs after Tmax (avoids dragging AUC down)
    )
  )
)
```

---

## AUC integration methods

The default integration method is **lin up / log down** (linear interpolation on the ascending phase, log-linear on the descending phase). Three methods are available:

```{r}
# Method 1: linear trapezoidal (simplest)
o_data_lin <- PKNCAdata(o_conc, o_dose, intervals = ev_intervals,
                         options = list(auc.method = "linear"))

# Method 2: lin-log (log trapezoidal throughout — not recommended)
o_data_linlog <- PKNCAdata(o_conc, o_dose, intervals = ev_intervals,
                            options = list(auc.method = "lin-log"))

# Method 3: lin up/log down (default, recommended)
o_data_liuplogdown <- PKNCAdata(o_conc, o_dose, intervals = ev_intervals,
                                 options = list(auc.method = "lin up/log down"))

# Compare AUClast across methods
compare_methods <- function(label, data_obj) {
  nca <- pk.nca(data_obj)
  as.data.frame(nca) |>
    filter(PPTESTCD == "auclast") |>
    mutate(method = label) |>
    select(method, subject, PPORRES)
}

bind_rows(
  compare_methods("linear",           o_data_lin),
  compare_methods("lin up/log down",  o_data_liuplogdown)
) |>
  tidyr::pivot_wider(names_from = method, values_from = PPORRES) |>
  arrange(subject)
```

---

## Units

```{r}
units_table <- pknca_units_table(
  concu   = "mg/L",
  timeu   = "h",
  doseu   = "mg",
  amountu = "mg"
)

o_conc_u <- PKNCAconc(d_conc, conc ~ time | subject, concu = "mg/L", timeu = "h")
o_dose_u <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular",
                      doseu = "mg", timeu = "h")
o_data_u <- PKNCAdata(o_conc_u, o_dose_u, intervals = ev_intervals, units = units_table)
o_nca_u  <- pk.nca(o_data_u)

as.data.frame(o_nca_u) |>
  filter(PPTESTCD %in% c("auclast", "aucinf.obs", "cl.obs", "half.life")) |>
  select(subject, PPTESTCD, PPORRES) |>
  arrange(subject, PPTESTCD)
```

---

## Summary

```{r}
summary(o_nca)
```

Customize for specific parameters:

```{r}
PKNCA.set.summary(
  "tmax",
  description = "median [min, max]",
  point  = median,
  spread = function(x) c(min(x), max(x))  # must return numeric; PKNCA formats the string
)

summary(o_nca)
```

---

## Full parameter list for extravascular

```{r}
interval_cols <- get.interval.cols()
ev_relevant <- c(
  "cmax", "tmax", "cmin", "tmin",
  "auclast", "aucinf.obs", "aucinf.pred",
  "aucpext.obs", "aucpext.pred",
  "half.life", "lambda.z", "lambda.z.n.points",
  "r.squared", "adj.r.squared",
  "cl.obs", "cl.pred", "vz.obs", "vz.pred",
  "mrt.last", "mrt.obs",
  "tlast", "tfirst", "clast.obs"
)

names(interval_cols)[names(interval_cols) %in% ev_relevant]
```
