---
title: "Half-Life and Terminal Phase"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)

d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)
o_conc <- PKNCAconc(d_conc, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")
```

## How PKNCA estimates λz

PKNCA uses **curve stripping**: it systematically evaluates all possible subsets of terminal-phase points (from 3 points up to the full tail), fits log-linear regression to each subset, and selects the best fit based on adjusted R².

The selection criterion uses a **tolerance window**: among all subsets that pass quality filters, PKNCA picks the one with the **most points** whose adjusted R² is within `adj.r.squared.factor` of the best adjusted R²:

$$\text{selected} = \arg\max_{\text{n points}} \left\{ \text{adj.}R^2 \geq \max(\text{adj.}R^2) - \text{adj.r.squared.factor} \right\}$$

The small tolerance (`adj.r.squared.factor`, default 0.0001) means: prefer longer windows (more points → more accurate λz) unless the longer window's adj.R² is noticeably worse than the best. This is a **model-selection tolerance**, not a per-point penalty added to R².

```{mermaid}
flowchart TD
    A["All points after Tmax"] --> B["Try subsets: last 3, last 4, ..., all"]
    B --> C["Fit log-linear regression to each"]
    C --> D{"Passes quality filters?\nλz > 0, R² ≥ threshold\nspan.ratio ≥ minimum"}
    D -- yes --> E["Compute adj.R²"]
    D -- no --> F["Discard subset"]
    E --> G["Select longest window within adj.r.squared.factor of best adj.R²"]
    G --> H["Return λz, half.life, R², n.points"]
```

---

## Basic half-life output

```{r}
hl_interval <- data.frame(
  start             = 0,
  end               = Inf,
  half.life         = TRUE,
  lambda.z          = TRUE,
  lambda.z.n.points = TRUE,
  lambda.z.time.first = TRUE,
  r.squared         = TRUE,
  adj.r.squared     = TRUE,
  span.ratio        = TRUE
)

o_nca <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = hl_interval))

as.data.frame(o_nca) |>
  filter(PPTESTCD %in% c("half.life", "lambda.z", "lambda.z.n.points",
                          "lambda.z.time.first", "adj.r.squared", "span.ratio")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

**What each output means:**

| Parameter | Meaning |
|---|---|
| `lambda.z` | Terminal elimination rate constant (h⁻¹) |
| `half.life` | ln(2) / λz |
| `lambda.z.n.points` | Number of points used in the selected regression |
| `lambda.z.time.first` | First timepoint included in the regression window |
| `r.squared` / `adj.r.squared` | Fit quality of the selected regression |
| `span.ratio` | (lambda.z.time.last − lambda.z.time.first) / half.life — post-hoc quality flag |

---

## Quality filters

PKNCA applies one filter **during fitting** and exposes additional parameters you use to flag or exclude results **after fitting** with `exclude()`.

> **Important distinction:** Only `min.hl.points` is enforced inside `pk.calc.half.life()` — too few points returns `half.life = NA` immediately. `min.hl.r.squared` and `min.span.ratio` are **not** consulted during fitting; they are thresholds for `exclude()` calls after `pk.nca()`.

### 1. Minimum number of points (enforced during fitting)

```{r}
# Default: 3 points minimum — the only filter that makes half.life NA during pk.nca()
o_data_4pts <- PKNCAdata(o_conc, o_dose, intervals = hl_interval,
                          options = list(min.hl.points = 4))
o_nca_4pts  <- pk.nca(o_data_4pts)

as.data.frame(o_nca_4pts) |>
  filter(PPTESTCD %in% c("half.life", "lambda.z.n.points")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

### 2. R² threshold (post-hoc exclusion via `exclude()`)

`min.hl.r.squared` sets the minimum **unadjusted R²** threshold. It has no effect during `pk.nca()`; use it with `exclude_nca_min.hl.r.squared()` after fitting:

```{r}
# Compute half-life for all subjects
o_nca_r2 <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = hl_interval))

# Then flag results with unadjusted R² < 0.95
o_nca_r2_excl <- exclude(
  o_nca_r2,
  reason = "r.squared < 0.95",
  FUN    = exclude_nca_min.hl.r.squared(0.95)
)

as.data.frame(o_nca_r2_excl) |>
  filter(PPTESTCD %in% c("half.life", "r.squared")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

For the stricter **adjusted R²** filter: `exclude_nca_min.hl.adj.r.squared(min)`.

### 3. Span ratio threshold (post-hoc exclusion via `exclude()`)

`min.span.ratio` has no effect during `pk.nca()`; use it with `exclude_nca_span.ratio()`:

```{r}
# Flag half-lives with span.ratio < 2
o_nca_span_excl <- exclude(
  o_nca_r2,
  reason = "span.ratio < 2",
  FUN    = exclude_nca_span.ratio(2)
)

as.data.frame(o_nca_span_excl) |>
  filter(PPTESTCD %in% c("half.life", "span.ratio")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

### 4. Allow / exclude Tmax

By default Tmax is excluded from the λz regression (drug may still be distributing).

```{r}
o_data_tmax <- PKNCAdata(o_conc, o_dose, intervals = hl_interval,
                          options = list(allow.tmax.in.half.life = TRUE))
o_nca_tmax  <- pk.nca(o_data_tmax)

as.data.frame(o_nca_tmax) |>
  filter(PPTESTCD %in% c("half.life", "lambda.z.n.points", "lambda.z.time.first")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

---

## Manually controlling which points are used

### Excluding specific points (`exclude_half.life`)

Mark individual observations that should never be included in the λz regression — for example, a suspected outlier or a resurgence peak.

```{r}
d_conc_excl <- d_conc |>
  mutate(excl_hl = ifelse(subject == "1" & time > 16, TRUE, NA))

o_conc_excl <- PKNCAconc(d_conc_excl, conc ~ time | subject,
                          exclude_half.life = "excl_hl")

o_nca_excl <- pk.nca(PKNCAdata(o_conc_excl, o_dose, intervals = hl_interval))

# Compare n.points and lambda.z.time.first for subject 1
bind_rows(
  as.data.frame(o_nca) |>
    filter(subject == "1", PPTESTCD %in% c("half.life", "lambda.z.n.points")) |>
    mutate(version = "default"),
  as.data.frame(o_nca_excl) |>
    filter(subject == "1", PPTESTCD %in% c("half.life", "lambda.z.n.points")) |>
    mutate(version = "late points excluded")
) |>
  select(version, PPTESTCD, PPORRES) |>
  arrange(version, PPTESTCD)
```

### Forcing specific points (`include_half.life`)

Mark points that **must** be included — they cannot be dropped even if excluding them would improve the fit. Useful for regulatory requirements.

```{r}
d_conc_incl <- d_conc |>
  mutate(incl_hl = ifelse(subject == "1" & time %in% c(9.05, 12.12), TRUE, NA))

o_conc_incl <- PKNCAconc(d_conc_incl, conc ~ time | subject,
                          include_half.life = "incl_hl")

o_nca_incl <- pk.nca(PKNCAdata(o_conc_incl, o_dose, intervals = hl_interval))

as.data.frame(o_nca_incl) |>
  filter(subject == "1",
         PPTESTCD %in% c("half.life", "lambda.z.n.points", "lambda.z.time.first")) |>
  select(PPTESTCD, PPORRES)
```

> **Note:** `include_half.life` forces the curve-stripping algorithm to always start at or before the earliest forced point. If the forced point is early in the profile, this can result in a worse fit (lower R²) — the point is still included even if it pulls the regression down.

---

## Tobit regression for BLQ-heavy data

Available in PKNCA ≥ 0.12.2 (now on CRAN).

### Why Tobit?

In studies where drug concentrations in the terminal elimination phase are frequently below the lower limit of quantification (BLQ), ordinary least-squares (OLS) log-linear regression produces biased estimates of λz. The bias arises because OLS either:

- **drops** BLQ observations entirely — shrinking the regression window and losing information about the true terminal slope, or
- **substitutes** BLQ = 0 — which is undefined on the log scale and forces arbitrary imputation.

Tobit regression handles this correctly by treating each BLQ observation as a **left-censored** data point. Rather than requiring a specific concentration value, it only asserts `C ≤ LLOQ`. The regression uses the full profile, including censored observations, without fabricating values.

### The key statistical difference from OLS

OLS minimises the sum of squared residuals across observed concentrations. Tobit instead **maximises a likelihood** that has two components:

1. **Observed concentrations** (above LLOQ): contribute via the normal log-likelihood of the log-transformed concentration given the fitted log-linear line.
2. **BLQ observations**: contribute the cumulative probability that the true (unobserved) concentration falls at or below the LLOQ, i.e. `Φ((log(LLOQ) − predicted) / σ)`.

This means BLQ timepoints actively inform the slope estimate — they pull the λz estimate toward the true terminal slope — rather than being discarded.

### When to consider Tobit

Tobit regression is most beneficial when:

- More than approximately **20% of terminal-phase observations are BLQ** — below this threshold, OLS and Tobit estimates converge.
- The drug has **low tissue concentrations** or a short distribution half-life relative to the sampling schedule, making late timepoints frequently BLQ.
- **Pediatric or geriatric studies** with limited sampling windows, where sparse late-time samples are often at or near the LLOQ.
- The terminal slope is important for downstream parameters (AUC extrapolation, mean residence time) and precision matters for regulatory submissions.

### Using Tobit (PKNCA ≥ 0.12.2)

Tobit is available by calling `pk.calc.half.life()` directly. Supply the full concentration-time vector and the LLOQ for each timepoint:

```{r}
# Simulated profile — concentrations at 16h and 24h are below the LLOQ of 0.1 mg/L
conc_tobit <- c(0, 2.5, 4.8, 4.2, 2.9, 1.4, 0.6, 0.05, 0.01)
time_tobit <- c(0, 0.5,   1,   2,   4,   8,  12,  16,  24)
lloq_tobit <- rep(0.1, length(conc_tobit))   # scalar lloq applied to all timepoints

# OLS reference
hl_ols   <- pk.calc.half.life(conc = conc_tobit, time = time_tobit, tmax = 1,
                               tlast = 24, hl_method = "log-linear")

# Tobit — includes the BLQ observations as left-censored at lloq
hl_tobit <- pk.calc.half.life(conc = conc_tobit, time = time_tobit, tmax = 1,
                               tlast = 24, lloq = lloq_tobit, hl_method = "tobit")

# Compare
data.frame(
  method    = c("log-linear", "tobit"),
  lambda.z  = c(hl_ols$lambda.z,  hl_tobit$lambda.z),
  half.life = c(hl_ols$half.life, hl_tobit$half.life),
  n.points  = c(hl_ols$lambda.z.n.points, hl_tobit$lambda.z.n.points)
)
```

> **Note:** Passing `hl_method = "tobit"` through the `pk.nca()` pipeline (via `PKNCAdata(options=...)`) also requires the LLOQ to be wired through from the concentration data, which is tracked for a future PKNCA update. For full pipeline support, use `hl_method = "log-linear"` (default) and call `pk.calc.half.life()` with `hl_method = "tobit"` directly for targeted analyses.

Tobit adds new output columns to the result:

| Parameter | Description |
|---|---|
| `lambda.z.n.points_blq` | Number of BLQ points included in the Tobit regression |
| `tobit_residual` | Sum of squared Tobit residuals for the selected window |
| `adj_tobit_residual` | Adjusted Tobit residual (analogous to adjusted R²) |

### Tobit-related options

| Option | Default | Effect |
|---|---|---|
| `hl_method` | `"log-linear"` | `"log-linear"` (OLS) or `"tobit"` (censored-likelihood regression) |
| `tobit_n_points_penalty` | `0` | Penalty per extra terminal point in the log-likelihood; positive values favour shorter windows |
| `tobit_optim_control` | `list()` | Control list forwarded to `stats::optim()` for the Tobit optimisation |

See `vignette("v06-half-life-calculation-tobit", package = "PKNCA")` for a detailed simulation comparison between OLS and Tobit estimates.

---

## lambda.z.corrxy — regression correlation

`lambda.z.corrxy` (available in PKNCA ≥ 0.12.2) returns the Pearson correlation between time and log-concentration for the points used in the λz regression. It is a unitless diagnostic — values close to −1 indicate a strong, clean terminal slope.

```{r}
hl_corrxy_interval <- data.frame(
  start              = 0,
  end                = Inf,
  half.life          = TRUE,
  lambda.z           = TRUE,
  lambda.z.corrxy    = TRUE,
  r.squared          = TRUE,
  adj.r.squared      = TRUE
)

o_nca_corrxy <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = hl_corrxy_interval))

as.data.frame(o_nca_corrxy) |>
  filter(PPTESTCD %in% c("half.life", "lambda.z.corrxy", "r.squared", "adj.r.squared")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

---

## Visualising the chosen regression window

```{r}
# Extract lambda.z.time.first and n.points to annotate the plot
hl_results <- as.data.frame(o_nca) |>
  filter(PPTESTCD %in% c("lambda.z.time.first", "lambda.z.n.points", "tmax")) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Plot subject 1 with regression window highlighted
subj1 <- d_conc |> filter(subject == "1")
win   <- hl_results |> filter(subject == "1")
start_t <- win$lambda.z.time.first

ggplot(subj1, aes(x = time, y = conc)) +
  geom_line(colour = "grey60") +
  geom_point(aes(colour = time >= start_t), size = 3) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
                      labels = c("excluded from λz", "used in λz"),
                      name = NULL) +
  scale_y_log10() +
  labs(title = "Subject 1 — terminal regression window (log scale)",
       x = "Time (h)", y = "Concentration (mg/L) [log scale]") +
  theme_minimal()
```


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html) · [pk.calc.half.life()](https://humanpred.github.io/pknca/reference/pk.calc.half.life.html) · [get_halflife_points()](https://humanpred.github.io/pknca/reference/get_halflife_points.html)
:::
