---
title: "Interval Selection and Parameter Catalogue"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

Intervals tell PKNCA *when* to calculate and *what* to calculate. Every NCA parameter is associated with exactly one interval row, and each row is independent of all others.

---

## 1. The Interval Data Frame Structure

An interval specification is a plain R `data.frame` with one row per calculation window. Two columns are required:

| Column  | Type    | Meaning                                              |
|---------|---------|------------------------------------------------------|
| `start` | numeric | Start of the calculation window (same units as time) |
| `end`   | numeric | End of the calculation window; use `Inf` for open-ended windows |

Every remaining column is a **parameter flag**: its name must match a known PKNCA parameter name and its value must be `TRUE` (calculate) or `FALSE` (skip). Columns set to `FALSE` are simply ignored, so you only need to list the parameters you actually want.

```{r interval-structure}
# A minimal interval: AUClast over the first 24 hours
my_intervals <- data.frame(
  start   = 0,
  end     = 24,
  auclast = TRUE
)
my_intervals
```

You can also request multiple parameters in a single row:

```{r interval-multi-param}
data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  half.life  = TRUE,
  aucinf.obs = TRUE
)
```

Column names cannot overlap with `start`, `end`, or any grouping variable name you use in `PKNCAconc()`.

---

## 2. Group Matching

PKNCA matches interval rows to concentration data by comparing the **shared column names** between the interval data frame and the grouping columns of the concentration object. Any column that appears in both is used as a matching key.

```{r group-matching}
data("Theoph")

d_conc <- PKNCAconc(Theoph, conc ~ Time | Subject)

# Intervals with a Subject column — each row applies to exactly one subject
subject_intervals <- data.frame(
  Subject = c("1", "2"),
  start   = c(0, 0),
  end     = c(24, 24),
  auclast = TRUE
)

# Intervals without Subject — this row applies to ALL subjects (default interval)
default_interval <- data.frame(
  start   = 0,
  end     = 24,
  auclast = TRUE
)
```

If an interval row has no group columns, it is a **default interval** and is applied to every group present in the concentration data that does not have a more specific match. This is useful when the same calculation windows apply to all subjects and analytes.

---

## 3. Default Intervals via `PKNCAdata()` with No `intervals` Argument

When you call `PKNCAdata()` without an explicit `intervals` argument, PKNCA substitutes the built-in default single-dose intervals. These are stored in `PKNCA.options("single.dose.aucs")`:

```{r default-intervals}
PKNCA.options("single.dose.aucs")
```

The two default rows are:

- **Row 1** (`end = 24`): calculates `auclast` only — the workhorse for most single-dose studies.
- **Row 2** (`end = Inf`): calculates `cmax`, `tmax`, `half.life`, and `aucinf.obs`.

These defaults are suitable for a single-dose extravascular study with at least 24 hours of sampling. For any other design (multiple dose, IV, urine, etc.) you should supply explicit intervals.

```{r use-default}
d_dose <- PKNCAdose(
  Theoph |> group_by(Subject) |>
    summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>
    mutate(Time = 0),
  Dose ~ Time | Subject
)

# No intervals argument — uses PKNCA.options("single.dose.aucs") automatically
d_data_default <- PKNCAdata(
  PKNCAconc(Theoph, conc ~ Time | Subject),
  d_dose
)
d_data_default$intervals
```

---

## 4. Multiple Intervals per Group

Each row of the interval data frame is applied **independently**. A single subject (or group) can have any number of interval rows, each with its own time window and parameter set. PKNCA applies every matching row and collects the results.

```{r multiple-intervals}
# Two windows for the same subjects:
# 0–24 h for early exposure; 0–Inf for terminal parameters
multi_intervals <- data.frame(
  start      = c(0, 0),
  end        = c(24, Inf),
  auclast    = c(TRUE,  FALSE),
  cmax       = c(FALSE, TRUE),
  half.life  = c(FALSE, TRUE),
  aucinf.obs = c(FALSE, TRUE)
)
multi_intervals
```

PKNCA processes each row separately: the first row restricts data to $t \in [0, 24]$ for the `auclast` calculation, while the second row uses all available data to derive terminal parameters.

Overlapping intervals are allowed and do not interfere with each other.

---

## 5. Duration-Based Events

Some sampling designs record a *duration* rather than an instantaneous sample — for example, urine collection periods or IV infusions. PKNCA tracks these via the `duration` argument in `PKNCAconc()` and `PKNCAdose()`.

When a concentration object carries duration information, PKNCA selects samples for an interval based on **both the start and end of the collection period**. The collection end time must fall at or before the interval `end` for the sample to be included:

```{r duration-concept, eval=FALSE}
# Urine example: each row covers a collection period
urine_conc <- data.frame(
  Subject  = rep(1, 4),
  Time     = c(0,  4,  8,  24),   # collection start times
  duration = c(4,  4,  16, 0),    # collection lengths (hours)
  conc     = c(0.5, 1.2, 0.9, 0.3)
)

d_urine_conc <- PKNCAconc(
  urine_conc,
  conc ~ Time | Subject,
  duration = "duration"
)

# The interval end must align with a collection end time
urine_intervals <- data.frame(
  start    = 0,
  end      = 24,   # matches the end of the 8-h collection (8 + 16 = 24)
  ae       = TRUE,
  clr.last = TRUE
)
```

For IV infusions, `setDuration()` on the dose object sets the infusion duration, and `ceoi` (concentration at end of infusion) becomes calculable within an interval whose `end` matches the infusion end time.

---

## 6. Full Parameter Catalogue

The complete list of calculable parameters is available programmatically via `get.interval.cols()`. Each entry carries a description, the calculation function, unit type, and dependency information.

```{r param-catalogue}
cols <- get.interval.cols()

param_table <- data.frame(
  parameter   = names(cols),
  description = sapply(cols, function(x) x$desc),
  stringsAsFactors = FALSE
)

# Exclude the structural columns start/end from the display
param_table <- param_table[!param_table$parameter %in% c("start", "end"), ]
rownames(param_table) <- NULL

knitr::kable(
  param_table,
  col.names = c("Parameter", "Description"),
  caption   = paste0(
    "All ", nrow(param_table),
    " calculable parameters in PKNCA ", packageVersion("PKNCA")
  )
)
```

Parameters fall into broad families:

- **AUC / AUMC** — `auclast`, `aucall`, `aucinf.obs`, `aucinf.pred`, `aumclast`, and interval-restricted `aucint.*` variants
- **Concentration** — `cmax`, `cmin`, `ctrough`, `clast.obs`, `clast.pred`, `c0`, `cstart`, `ceoi`
- **Time** — `tmax`, `tlast`, `tfirst`, `tlag`
- **Clearance / Volume** — `cl.obs`, `cl.pred`, `vz.obs`, `vz.pred`, `vss.obs`, `vss.iv.obs`, renal `clr.*`
- **Elimination** — `half.life`, `lambda.z`, `lambda.z.n.points`, `span.ratio`, `r.squared`, `adj.r.squared`
- **Excretion** — `ae`, `fe`
- **Steady-state / MD** — `deg.fluc`, `swing`, `ptr`, `cav`, `mrt.*`, `thalf.eff.*`
- **Sparse sampling** — `sparse_auclast`, `sparse_auc_se`, `sparse_auc_df`
- **IV-specific** — `aucivlast`, `aucivpbext*`, `aucivinf.*`
- **Dose-normalized** — `auclast.dn`, `cmax.dn`, and other `*.dn` variants

---

## 7. Checking Parameter Dependencies with `get.parameter.deps()`

Some parameters depend on others being computed first. `get.parameter.deps()` returns all parameters that use a given parameter as an input:

```{r param-deps}
# Which parameters depend on auclast?
get.parameter.deps("auclast")
```

This is useful when you enable a high-level parameter and want to know which intermediate quantities will also be computed. Dependencies are resolved automatically — you do not need to set dependency flags to `TRUE` manually.

---

## 8. Validating Custom Intervals with `check.interval.specification()`

Before passing intervals to `PKNCAdata()`, you can validate them with `check.interval.specification()`. It verifies that all column names are recognized parameters, that `start < end`, and that logical values are used correctly.

```{r check-intervals}
# A valid interval
valid_int <- data.frame(
  start      = c(0, 0),
  end        = c(24, Inf),
  auclast    = c(TRUE, FALSE),
  aucinf.obs = c(FALSE, TRUE),
  cmax       = c(FALSE, TRUE)
)
check.interval.specification(valid_int)
```

```{r check-invalid, error=TRUE}
# An invalid interval — start >= end
bad_int <- data.frame(
  start   = 24,
  end     = 0,
  auclast = TRUE
)
check.interval.specification(bad_int)
```

`check.interval.specification()` returns the interval data frame invisibly on success, making it safe to use inline before passing to `PKNCAdata()`:

```{r check-inline}
d_conc <- PKNCAconc(Theoph, conc ~ Time | Subject)
d_dose <- PKNCAdose(
  Theoph |> group_by(Subject) |>
    summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>
    mutate(Time = 0),
  Dose ~ Time | Subject
)

my_intervals <- check.interval.specification(
  data.frame(
    start      = c(0, 0),
    end        = c(24, Inf),
    auclast    = c(TRUE, FALSE),
    aucinf.obs = c(FALSE, TRUE),
    cmax       = c(FALSE, TRUE),
    half.life  = c(FALSE, TRUE)
  )
)

d_data <- PKNCAdata(d_conc, d_dose, intervals = my_intervals)
suppressMessages(o_nca <- pk.nca(d_data))
as.data.frame(o_nca) |> head(8)
```
