---
title: "Intravascular (IV) Examples"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)    # load after PKNCA so dplyr::filter wins
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## The IV dataset: Indomethacin

We use `datasets::Indometh` — IV bolus indomethacin in 6 subjects, sampled over 8 hours.

```{r}
head(Indometh)
str(Indometh)
```

Columns: `Subject` (factor, 6 levels), `time` (h), `conc` (mcg/mL).

We add dose information. The standard IV bolus dose for this study was 25 mg.

```{r}
d_conc <- as.data.frame(Indometh)

d_dose <- data.frame(
  Subject = unique(d_conc$Subject),
  dose    = 25,   # mg
  time    = 0,
  route   = "intravascular"
)

# Quick look at concentration profiles
ggplot(d_conc, aes(x = time, y = conc, group = Subject, colour = Subject)) +
  geom_line() + geom_point() +
  scale_y_log10() +
  labs(title = "Indomethacin IV — log-linear concentration profiles",
       x = "Time (h)", y = "Concentration (mcg/mL)") +
  theme_minimal()
```

The log-linear profiles confirm monoexponential decline — well suited for IV NCA.

---

## Basic IV analysis

```{r}
o_conc <- PKNCAconc(d_conc, conc ~ time | Subject)

o_dose <- PKNCAdose(d_dose, dose ~ time | Subject, route = "intravascular")

# Indometh: no sample at t = 0 (first sample is at 0.25h).
# Use impute = "start_conc0" to add conc = 0 at t = 0 so the interval can start at 0.
# This slightly underestimates AUClast vs. true C0 back-extrapolation (see C0 section below).
first_t <- 0  # interval starts at 0 with imputed C0 = 0

# Request the core IV parameters
iv_intervals <- data.frame(
  start       = first_t,
  end         = Inf,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE,
  cmax        = TRUE,
  tmax        = TRUE,
  lambda.z    = TRUE,
  cl.obs      = TRUE,
  vz.obs      = TRUE,
  mrt.iv.last = TRUE
)

o_data <- PKNCAdata(o_conc, o_dose, intervals = iv_intervals, impute = "start_conc0")
o_nca  <- pk.nca(o_data)

as.data.frame(o_nca) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

---

## Parameter reference: what each one means

### Cmax and Tmax

For IV bolus, Cmax is the first measured concentration (no absorption phase).
Tmax is the time of that first sample.

```{r}
as.data.frame(o_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax")) |>
  select(Subject, PPTESTCD, PPORRES)
```

### C0 — extrapolated initial concentration

For IV bolus, the true concentration at time 0 is not directly measured (the first sample is taken minutes later). PKNCA extrapolates C0 by back-extrapolating the log-linear terminal phase.

```{r}
iv_with_c0 <- data.frame(
  start    = first_t,
  end      = Inf,
  c0       = TRUE,
  auclast  = TRUE,
  aucinf.obs = TRUE
)

o_data_c0 <- PKNCAdata(o_conc, o_dose, intervals = iv_with_c0, impute = "start_conc0")
o_nca_c0  <- pk.nca(o_data_c0)

as.data.frame(o_nca_c0) |>
  filter(PPTESTCD %in% c("c0", "cmax", "auclast", "aucinf.obs")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

> **Note:** C0 > Cmax is expected for IV bolus — Cmax is the first measured point, C0 is extrapolated back to time 0.

### AUC variants

| Parameter | Meaning |
|---|---|
| `auclast` | AUC from 0 to last measurable concentration (trapezoidal) |
| `aucinf.obs` | AUC extrapolated to ∞, using observed Clast |
| `aucinf.pred` | AUC extrapolated to ∞, using predicted Clast from λz fit |
| `aucpext.obs` | % of AUCinf that is extrapolated (quality check) |

```{r}
auc_params <- data.frame(
  start       = first_t,
  end         = Inf,
  auclast     = TRUE,
  aucinf.obs  = TRUE,
  aucinf.pred = TRUE,
  aucpext.obs = TRUE
)

o_data_auc <- PKNCAdata(o_conc, o_dose, intervals = auc_params, impute = "start_conc0")
o_nca_auc  <- pk.nca(o_data_auc)

as.data.frame(o_nca_auc) |>
  filter(PPTESTCD %in% c("auclast", "aucinf.obs", "aucinf.pred", "aucpext.obs")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

> **Quality check:** `aucpext.obs` > 20% (the default `max.aucinf.pext` threshold) means too much of the AUC is extrapolated — the study may not have sampled long enough.

### Half-life and λz

λz (lambda.z) is the terminal elimination rate constant estimated from log-linear regression of the terminal phase. Half-life = ln(2) / λz.

```{r}
hl_params <- data.frame(
  start        = first_t,
  end          = Inf,
  lambda.z     = TRUE,
  half.life    = TRUE,
  r.squared    = TRUE,
  lambda.z.n.points = TRUE
)

o_data_hl <- PKNCAdata(o_conc, o_dose, intervals = hl_params, impute = "start_conc0")
o_nca_hl  <- pk.nca(o_data_hl)

as.data.frame(o_nca_hl) |>
  filter(PPTESTCD %in% c("lambda.z", "half.life", "r.squared", "lambda.z.n.points")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

**Controlling half-life quality:**

```{r}
# Stricter: require at least 4 points and R² ≥ 0.95
o_data_strict <- PKNCAdata(
  o_conc, o_dose,
  intervals = hl_params,
  impute = "start_conc0",
  options = list(
    min.hl.points    = 4,
    min.hl.r.squared = 0.95
  )
)
o_nca_strict <- pk.nca(o_data_strict)

as.data.frame(o_nca_strict) |>
  filter(PPTESTCD %in% c("half.life", "r.squared", "lambda.z.n.points")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

**Excluding specific points from λz:**

Use `exclude_half.life` column in your concentration data to mark points that should not be included in the terminal regression:

```{r}
d_conc_excl <- d_conc |>
  mutate(excl_hl = ifelse(Subject == "1" & time < 0.5, TRUE, NA))

o_conc_excl <- PKNCAconc(
  d_conc_excl,
  conc ~ time | Subject,
  exclude_half.life = "excl_hl"
)
```

### Clearance and Volume

For IV, apparent clearance and volume can be calculated directly (no bioavailability factor needed).

| Parameter | Formula | Meaning |
|---|---|---|
| `cl.obs` | Dose / AUCinf.obs | Total body clearance (L/h) |
| `cl.pred` | Dose / AUCinf.pred | CL using predicted Clast |
| `vz.obs` | CL / λz | Volume of distribution (L) |
| `vz.pred` | CL.pred / λz | Vz using predicted Clast |
| `vss.obs` | CL × MRT | Volume at steady state |

```{r}
cl_params <- data.frame(
  start    = first_t,
  end      = Inf,
  cl.obs   = TRUE,
  cl.pred  = TRUE,
  vz.obs   = TRUE,
  vz.pred  = TRUE
)

o_data_cl <- PKNCAdata(o_conc, o_dose, intervals = cl_params, impute = "start_conc0")
o_nca_cl  <- pk.nca(o_data_cl)

as.data.frame(o_nca_cl) |>
  filter(PPTESTCD %in% c("cl.obs", "cl.pred", "vz.obs", "vz.pred")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

### Mean Residence Time (MRT)

MRT quantifies the average time a drug molecule spends in the body.

For IV, the AUMC (area under the first moment curve) method gives:

```{r}
mrt_params <- data.frame(
  start       = first_t,
  end         = Inf,
  mrt.iv.last = TRUE,
  mrt.iv.inf.obs = TRUE
)

o_data_mrt <- PKNCAdata(o_conc, o_dose, intervals = mrt_params, impute = "start_conc0")
o_nca_mrt  <- pk.nca(o_data_mrt)

as.data.frame(o_nca_mrt) |>
  filter(grepl("^mrt", PPTESTCD)) |>
  select(Subject, PPTESTCD, PPORRES)
```

---

## IV infusion (non-bolus)

For IV infusions, supply `duration` (infusion duration in time units) or `rate` in `PKNCAdose`:

```{r}
# Simulate: same dose but given as 30-minute infusion
d_dose_inf <- d_dose |>
  mutate(duration = 0.5)  # 0.5 hours = 30 min

o_dose_inf <- PKNCAdose(
  d_dose_inf,
  dose ~ time | Subject,
  route    = "intravascular",
  duration = "duration"   # column name
)
```

> For infusions, PKNCA adjusts the MRT calculation: MRT_infusion = MRT_bolus + Duration/2.

---

## AUC over a specific interval

For steady-state or partial AUC windows:

```{r}
# AUC from first sample to 4 hours
partial_auc <- data.frame(
  start   = first_t,
  end     = 4,
  auclast = TRUE,
  cmax    = TRUE
)

o_data_partial <- PKNCAdata(o_conc, o_dose, intervals = partial_auc, impute = "start_conc0")
o_nca_partial  <- pk.nca(o_data_partial)

as.data.frame(o_nca_partial) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject)
```

---

## Units

Assign units to get automatic unit propagation through derived parameters:

```{r}
units_table <- pknca_units_table(
  concu  = "mcg/mL",
  timeu  = "h",
  doseu  = "mg",
  amountu = "mg"
)

o_conc_u <- PKNCAconc(d_conc, conc ~ time | Subject, concu = "mcg/mL", timeu = "h")
o_dose_u <- PKNCAdose(d_dose, dose ~ time | Subject, route = "intravascular",
                      doseu = "mg", timeu = "h")
o_data_u <- PKNCAdata(o_conc_u, o_dose_u, intervals = iv_intervals, units = units_table,
                     impute = "start_conc0")
o_nca_u  <- pk.nca(o_data_u)

as.data.frame(o_nca_u) |>
  filter(PPTESTCD %in% c("auclast", "cl.obs", "vz.obs")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

---

## Excluding data points

Mark individual observations to exclude from calculations (e.g., suspected sampling errors):

```{r}
d_conc_with_excl <- d_conc |>
  mutate(excl = ifelse(Subject == "2" & time == 2, "suspected contamination", NA_character_))

o_conc_excl <- PKNCAconc(d_conc_with_excl, conc ~ time | Subject, exclude = "excl")
```

Excluded points are tracked in the results and appear in `as.data.frame()` output.

---

## Summary with custom statistics

```{r}
# Default summary uses geometric mean ± CV% for most parameters
summary(o_nca)
```

Customize the summary function (e.g., arithmetic mean ± SD for Tmax):

```{r}
PKNCA.set.summary(
  "tmax",
  description = "median [range]",
  point = median,
  spread = function(x) paste0("[", min(x), ", ", max(x), "]")
)

summary(o_nca)
```

---

## Full parameter list for IV

Below are all IV-relevant parameters you can request in an interval:

```{r}
# Show all available interval columns
interval_cols <- get.interval.cols()
iv_relevant <- c(
  "auclast", "aucinf.obs", "aucinf.pred", "aucpext.obs", "aucpext.pred",
  "aumc.last", "aumc.inf.obs",
  "c0", "cmax", "tmax", "tlast", "clast.obs",
  "half.life", "lambda.z", "lambda.z.n.points", "r.squared", "adj.r.squared",
  "cl.obs", "cl.pred", "vz.obs", "vz.pred",
  "mrt.iv.last", "mrt.iv.inf.obs", "mrt.iv.inf.pred"
)

# Which of these are registered in PKNCA?
names(interval_cols)[names(interval_cols) %in% iv_relevant]
```
