---
title: "Multiple-Dose and Steady-State"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## Multiple-dose NCA concepts

At steady state, the dosing interval τ (tau) is the key unit of analysis. Instead of analyzing from 0 to ∞, you analyze within one dosing interval [0, τ]. Additional parameters describe how concentrations fluctuate over the interval.

| Term | Meaning |
|---|---|
| τ (tau) | Dosing interval (h) |
| Cmin / Ctrough | Concentration at end of τ (= start of next dose) |
| Cmax | Peak concentration within τ |
| Cav | Average concentration over τ = AUCτ / τ |
| Peak-trough ratio (PTR) | Cmax / Cmin |
| Degree of fluctuation | (Cmax − Cmin) / Cav |
| Swing | (Cmax − Cmin) / Cmin |

---

## Synthetic steady-state dataset

We simulate steady-state oral theophylline to demonstrate all parameters. The key is that `start` and `end` define a single dosing interval.

```{r}
set.seed(42)

# Theoph-like PK: Ka=1.5/h, ke=0.08/h, V=30L, dose=400mg
ka <- 1.5; ke <- 0.08; V <- 30; dose <- 400; tau <- 12

# One-compartment oral steady-state: C(t) = [dose/V * ka/(ka-ke)] * [e^(-ke*t)/(1-e^(-ke*tau)) - e^(-ka*t)/(1-e^(-ka*tau))]
C_ss <- function(t, ka, ke, V, dose, tau) {
  A <- (dose / V) * (ka / (ka - ke))
  A * (exp(-ke * t) / (1 - exp(-ke * tau)) - exp(-ka * t) / (1 - exp(-ka * tau)))
}

timepoints <- c(0, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)
n_subjects <- 4

d_conc_ss <- expand.grid(Subject = factor(1:n_subjects), time = timepoints) |>
  mutate(
    # Add ~20% log-normal IIV on clearance (ke)
    ke_i  = ke * exp(rnorm(n(), 0, 0.2)),
    conc  = mapply(C_ss, time, ka, ke_i, V, dose, tau) + rnorm(n(), 0, 0.05),
    conc  = pmax(conc, 0)
  ) |>
  select(Subject, time, conc)

d_dose_ss <- data.frame(
  Subject = factor(1:n_subjects),
  dose    = dose,
  time    = 0,
  route   = "extravascular"
)

ggplot(d_conc_ss, aes(x = time, y = conc, group = Subject, colour = Subject)) +
  geom_line() + geom_point() +
  labs(title = "Simulated steady-state oral profiles (one dosing interval)",
       x = "Time within interval (h)", y = "Concentration (mg/L)") +
  theme_minimal()
```

---

## Basic steady-state NCA

For a single dosing interval, set `end = tau`. Parameters like `cav`, `ctrough`, `ptr`, `deg.fluc`, and `swing` all require a finite interval end.

```{r}
o_conc_ss <- PKNCAconc(d_conc_ss, conc ~ time | Subject)
o_dose_ss <- PKNCAdose(d_dose_ss, dose ~ time | Subject, route = "extravascular")

ss_intervals <- data.frame(
  start    = 0,
  end      = tau,
  auclast  = TRUE,
  cmax     = TRUE,
  tmax     = TRUE,
  ctrough  = TRUE,
  cstart   = TRUE,
  cav      = TRUE,
  ptr      = TRUE,
  deg.fluc = TRUE,
  swing    = TRUE
)

o_data_ss <- PKNCAdata(o_conc_ss, o_dose_ss, intervals = ss_intervals)
o_nca_ss  <- pk.nca(o_data_ss)

as.data.frame(o_nca_ss) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "ctrough", "cstart", "cav",
                          "ptr", "deg.fluc", "swing")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(Subject)
```

---

## Parameter reference

### Cstart and Ctrough

`cstart` is the concentration at the **start** of the interval (time = `start`). For steady state this equals the pre-dose trough.
`ctrough` is the concentration at the **end** of the interval (time = `end`). For steady state, `cstart` ≈ `ctrough`.

```{r}
as.data.frame(o_nca_ss) |>
  filter(PPTESTCD %in% c("cstart", "ctrough")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```

### Cav — average concentration

`cav = AUCτ / τ`. Useful for computing clearance at steady state (CL/F = dose / (AUCτ)).

```{r}
as.data.frame(o_nca_ss) |>
  filter(PPTESTCD %in% c("cav", "auclast")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(cav_check = auclast / tau) |>
  arrange(Subject)
```

### PTR, degree of fluctuation, swing

These describe how much concentrations vary within a dosing interval — important for characterizing exposure variability and clinical relevance.

| Parameter | Formula |
|---|---|
| `ptr` | Cmax / Ctrough |
| `deg.fluc` | (Cmax − Cmin) / Cav × 100% |
| `swing` | (Cmax − Cmin) / Cmin × 100% |

```{r}
as.data.frame(o_nca_ss) |>
  filter(PPTESTCD %in% c("ptr", "deg.fluc", "swing")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(Subject)
```

### Ceoi — concentration at end of infusion

For IV infusions, `ceoi` is the concentration at the moment the infusion ends (= `start + duration`). For bolus or oral dosing, it equals `cmax` at the appropriate time.

```{r}
ceoi_interval <- data.frame(
  start  = 0,
  end    = tau,
  ceoi   = TRUE,
  cmax   = TRUE
)

o_nca_ceoi <- pk.nca(PKNCAdata(o_conc_ss, o_dose_ss, intervals = ceoi_interval))

as.data.frame(o_nca_ceoi) |>
  filter(PPTESTCD %in% c("ceoi", "cmax")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(Subject)
```

---

## Multiple dose intervals in one dataset

When data span multiple dose occasions, define one interval row per occasion per group. PKNCA applies each interval independently.

```{r}
# Two dosing intervals: 0-12h and 12-24h (simulate second interval)
d_conc_2dose <- bind_rows(
  d_conc_ss |> mutate(occ = 1),
  d_conc_ss |> mutate(time = time + 12, occ = 2)
)

# Two dose events
d_dose_2dose <- bind_rows(
  d_dose_ss |> mutate(time = 0),
  d_dose_ss |> mutate(time = 12)
)

o_conc_2d <- PKNCAconc(d_conc_2dose, conc ~ time | Subject)
o_dose_2d <- PKNCAdose(d_dose_2dose, dose ~ time | Subject, route = "extravascular")

# One interval per occasion
two_dose_intervals <- data.frame(
  start    = c(0, 12),
  end      = c(12, 24),
  auclast  = TRUE,
  cmax     = TRUE,
  ctrough  = TRUE,
  deg.fluc = TRUE
)

o_nca_2d <- pk.nca(PKNCAdata(o_conc_2d, o_dose_2d, intervals = two_dose_intervals))

as.data.frame(o_nca_2d) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "ctrough")) |>
  select(Subject, start, end, PPTESTCD, PPORRES) |>
  arrange(Subject, start, PPTESTCD)
```

---

## Finding tau automatically: `find.tau()`

When the dosing interval is not known a priori, `find.tau()` estimates it from the dosing times in a `PKNCAdose` object.

```{r}
# find.tau returns the most common dosing interval
find.tau(o_dose_2d)
```

Use the result to construct intervals programmatically:

```{r}
tau_detected <- find.tau(o_dose_2d)
cat("Detected tau:", tau_detected, "h\n")
```

---

## Vss at multiple dose (`vss.md.*`)

`vss.md.obs` and `vss.md.pred` are volume of distribution at steady state computed using the multiple-dose MRT formula. They require `auclast` (over a tau interval) and dose.

```{r}
vss_md_interval <- data.frame(
  start      = 0,
  end        = tau,
  vss.md.obs = TRUE,
  vss.md.pred = TRUE
)

o_nca_vss <- pk.nca(PKNCAdata(o_conc_ss, o_dose_ss, intervals = vss_md_interval))

as.data.frame(o_nca_vss) |>
  filter(PPTESTCD %in% c("vss.md.obs", "vss.md.pred")) |>
  select(Subject, PPTESTCD, PPORRES) |>
  arrange(Subject, PPTESTCD)
```
