---
title: "Post-Processing Results"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)

d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], weight = Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)

o_conc <- PKNCAconc(d_conc, conc ~ time | subject)
o_dose <- PKNCAdose(d_dose, dose ~ time | subject, route = "extravascular")
o_data <- PKNCAdata(o_conc, o_dose, intervals = data.frame(
  start = 0, end = Inf,
  auclast = TRUE, aucinf.obs = TRUE, cmax = TRUE,
  half.life = TRUE, span.ratio = TRUE, adj.r.squared = TRUE,
  aucpext.obs = TRUE
))
o_nca <- pk.nca(o_data)
```

## Why post-process?

After `pk.nca()` you may need to:

1. **Exclude** specific results that failed QC (poor half-life fit, protocol deviation)
2. **Normalize** parameters by dose (built-in `.dn` variants)
3. **Summarize** with custom statistics, respecting exclusions

All of this operates on the `PKNCAresults` object without rerunning the analysis.

---

## Excluding results

### Manual exclusion with `exclude()`

`exclude()` takes your results object plus a `reason` string and either a logical `mask` or a rule `FUN`. Excluding a parameter automatically excludes all downstream parameters that depend on it.

```{r}
# Exclude subject 3's half-life manually (e.g., failed QC review)
o_nca_excl <- exclude(
  o_nca,
  reason = "half-life failed manual QC",
  mask   = as.data.frame(o_nca)$subject == "3" &
           as.data.frame(o_nca)$PPTESTCD == "half.life"
)

# Exclusion propagates to aucinf.obs (depends on lambda.z which depends on half.life)
as.data.frame(o_nca_excl) |>
  filter(subject == "3", !is.na(exclude)) |>
  select(subject, PPTESTCD, PPORRES, exclude)
```

### Rule-based exclusion functions

PKNCA provides ready-made rule functions to pass to `exclude(FUN=...)`:

| Function | Excludes when... |
|---|---|
| `exclude_nca_span.ratio(min)` | span.ratio < min |
| `exclude_nca_min.hl.r.squared(min)` | half-life adj.R² < min |
| `exclude_nca_min.hl.adj.r.squared(min)` | adjusted R² < min (stricter) |
| `exclude_nca_max.aucinf.pext(max)` | %AUCextrap > max |
| `exclude_nca_tmax_0()` | Tmax = 0 (no absorption observed) |
| `exclude_nca_tmax_early()` | Tmax is at first timepoint |
| `exclude_nca_count_conc_measured(min)` | too few measured concentrations |
| `exclude_nca_by_param(params)` | exclude specific parameters by name |

```{r}
# Exclude half-lives where span ratio < 2
o_nca_span <- exclude(
  o_nca,
  reason = "span.ratio < 2",
  FUN    = exclude_nca_span.ratio(2)
)

as.data.frame(o_nca_span) |>
  filter(PPTESTCD == "half.life") |>
  select(subject, PPORRES, exclude) |>
  arrange(subject)
```

```{r}
# Chain multiple exclusion rules
o_nca_multi <- o_nca |>
  exclude(reason = "span.ratio < 2",    FUN = exclude_nca_span.ratio(2)) |>
  exclude(reason = "%AUCextrap > 20%",  FUN = exclude_nca_max.aucinf.pext(20))

as.data.frame(o_nca_multi) |>
  filter(!is.na(exclude)) |>
  select(subject, PPTESTCD, exclude) |>
  arrange(subject, PPTESTCD)
```

### Exclusions are respected in summary

```{r}
summary(o_nca_span)
```

---

## Dose-normalized parameters

PKNCA has built-in dose-normalized variants for several parameters. Request them in the interval — they are automatically computed as `parameter / dose`.

```{r}
dn_interval <- data.frame(
  start        = 0,
  end          = Inf,
  auclast      = TRUE,
  auclast.dn   = TRUE,   # AUClast / dose
  aucinf.obs   = TRUE,
  aucinf.obs.dn = TRUE,  # AUCinf / dose
  cmax         = TRUE,
  cmax.dn      = TRUE    # Cmax / dose
)

o_nca_dn <- pk.nca(PKNCAdata(o_conc, o_dose, intervals = dn_interval))

as.data.frame(o_nca_dn) |>
  filter(PPTESTCD %in% c("auclast", "auclast.dn", "cmax", "cmax.dn")) |>
  select(subject, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(subject)
```

All available dose-normalized parameters:

```{r}
cols <- get.interval.cols()
dn_params <- names(cols)[endsWith(names(cols), ".dn")]
data.frame(
  parameter       = dn_params,
  normalized_from = sub("\\.dn$", "", dn_params)
) |> knitr::kable()
```

---

## Extracting results

`as.data.frame()` returns the full tidy result including excluded rows:

```{r}
df <- as.data.frame(o_nca)

# Columns:
# subject    — group identifier
# start/end  — interval boundaries
# PPTESTCD   — parameter code
# PPORRES    — numeric result
# exclude    — NA if valid; character reason if excluded

df |>
  filter(PPTESTCD == "auclast") |>
  select(subject, PPORRES, exclude) |>
  arrange(subject)
```

---

## Custom summary statistics

`PKNCA.set.summary()` controls how each parameter is presented in `summary()`.

```{r}
# Arithmetic mean ± SD for AUClast
PKNCA.set.summary(
  "auclast",
  description = "mean ± SD",
  point  = mean,
  spread = sd
)

# Median [min, max] for Tmax
# spread must return a 2-element numeric vector
PKNCA.set.summary(
  "tmax",
  description = "median [min, max]",
  point  = median,
  spread = function(x) c(min(x), max(x))
)

summary(o_nca)
```

Restore to default (geometric mean ± geometric CV%):

```{r}
PKNCA.set.summary(
  "auclast",
  description = "geometric mean [geometric CV%]",
  point  = PKNCA:::geomean,
  spread = PKNCA:::geocv
)

PKNCA.set.summary(
  "tmax",
  description = "median [min, max]",
  point  = median,
  spread = function(x) c(min(x), max(x))
)
```

---

## Business summary helpers

PKNCA exports two summary functions suitable for use in regulatory submission tables. They are also used internally when `PKNCA.set.summary()` is called with defaults.

| Function | Formula | Typical use |
|---|---|---|
| `business.geomean(x)` | exp(mean(log(x))) | Geometric mean |
| `business.geocv(x)` | sqrt(exp(var(log(x))) − 1) × 100 | Geometric coefficient of variation (%) |

```{r}
x <- c(10.2, 11.8, 9.6, 12.1, 10.5)
cat("Geometric mean:", business.geomean(x), "\n")
cat("Geometric CV%: ", business.geocv(x),   "\n")
```

These are useful when computing summary statistics outside of `summary()`, for example in custom reporting pipelines.

---

## Inspecting the terminal regression window

`get_halflife_points()` returns the individual observations that were included in the selected λz regression for each subject. Useful for QC review or plotting.

```{r}
d_conc_pp <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose_pp <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)

o_conc_pp <- PKNCAconc(d_conc_pp, conc ~ time | subject)
o_dose_pp <- PKNCAdose(d_dose_pp, dose ~ time | subject, route = "extravascular")
o_nca_pp  <- pk.nca(PKNCAdata(o_conc_pp, o_dose_pp,
                               intervals = data.frame(start=0, end=Inf, half.life=TRUE)))

# Get the points used in each subject's terminal regression
hl_pts <- get_halflife_points(o_nca_pp)
head(hl_pts)
```

```{r}
# Plot: highlight regression points for subject 1
subj1_conc <- d_conc_pp |> filter(subject == "1")
subj1_pts  <- hl_pts    |> filter(subject == "1")

ggplot(subj1_conc, aes(x = time, y = conc)) +
  geom_line(colour = "grey70") +
  geom_point(colour = "grey50", size = 2) +
  geom_point(data = subj1_pts, aes(x = time, y = conc),
             colour = "firebrick", size = 3, shape = 21, fill = "firebrick", alpha = 0.6) +
  scale_y_log10() +
  labs(title = "Subject 1 — points used in λz regression (red)",
       x = "Time (h)", y = "Concentration (mg/L) [log scale]") +
  theme_minimal()
```

---

## Imputation methods

PKNCA supports pre-dose and post-dose concentration imputation strategies, controlled by the `impute` argument to `PKNCAdata()` or by columns in the concentration data.

Built-in imputation strategies:

| Strategy | When it applies | What it does |
|---|---|---|
| `"start_conc0"` | Interval start before first measured time | Adds conc = 0 at the interval start time |
| `"start_predose"` | Interval start before first measured time | Uses the last pre-dose concentration |
| `"none"` | Any | No imputation — interval must start at or after first sample |

```{r}
# Default (no imputation) vs start_conc0
d_conc_imp <- d_conc_pp |> filter(subject == "1")
o_conc_imp <- PKNCAconc(d_conc_imp, conc ~ time | subject)
o_dose_imp <- PKNCAdose(
  d_dose_pp |> filter(subject == "1"),
  dose ~ time | subject,
  route = "extravascular"
)

# Without imputation: interval must start at/after first timepoint (0)
o_nca_noimpt <- pk.nca(PKNCAdata(o_conc_imp, o_dose_imp,
  intervals = data.frame(start = 0, end = Inf, auclast = TRUE)))

# With start_conc0: adds a 0 at t = 0 if not present
o_nca_impt <- pk.nca(PKNCAdata(o_conc_imp, o_dose_imp,
  intervals = data.frame(start = 0, end = Inf, auclast = TRUE),
  impute = "start_conc0"))

bind_rows(
  as.data.frame(o_nca_noimpt) |> filter(PPTESTCD=="auclast") |> mutate(impute="none"),
  as.data.frame(o_nca_impt)   |> filter(PPTESTCD=="auclast") |> mutate(impute="start_conc0")
) |>
  select(impute, PPTESTCD, PPORRES)
```
