---
title: "Regulatory and CDISC Context"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

PKNCA does not include a dedicated CDISC submission function. What it does provide is a tidy, long-format result table whose structure maps directly onto the CDISC PP (Pharmacokinetic Parameters) domain. This page documents that mapping and describes how to prepare outputs for regulatory submissions.

```{r build-result, include=FALSE}
data("Theoph")
d_conc <- PKNCAconc(Theoph, conc ~ Time | Subject)
d_dose <- PKNCAdose(
  Theoph |> group_by(Subject) |>
    summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>
    mutate(Time = 0),
  Dose ~ Time | Subject
)
d_data <- PKNCAdata(d_conc, d_dose)
suppressMessages(o_nca <- pk.nca(d_data))
```

---

## 1. CDISC PP Domain Alignment

The CDISC PP domain stores one parameter value per row, with the parameter code in `PPTESTCD`, the numeric result in `PPORRES`, and the standardized result in `PPSTRES` (character) when a units table is provided. PKNCA's `as.data.frame()` output follows exactly this long-format convention.

```{r result-columns}
df <- as.data.frame(o_nca)
names(df)
```

The columns map to CDISC PP variables as follows:

| PKNCA column | CDISC PP variable | Notes                                                  |
|--------------|-------------------|--------------------------------------------------------|
| `PPTESTCD`   | `PPTESTCD`        | Parameter short name; identical names by convention    |
| `PPORRES`    | `PPORRES`         | Original numeric result (character in CDISC; numeric here) |
| `start`      | —                 | Interval start time; useful for `PPRFTDTC` derivation |
| `end`        | —                 | Interval end time                                      |
| `exclude`    | `PPEXCLFL` / `PPDRVFL` | Non-`NA` string gives the exclusion reason        |
| Grouping columns (`Subject`, etc.) | `USUBJID`, `PPCAT` | Pass through unchanged |

Units (`PPORRESU`, `PPSTRESU`) are available when you configure a units table via `pknca_units_table()` and pass it to `PKNCAdata()`. Without a units table, `PPORRES` carries the raw numeric value and unit derivation must be done externally.

```{r result-head}
head(df, 10)
```

---

## 2. Grouping Columns as CDISC Identifiers

Whatever grouping structure you define in `PKNCAconc()` — subject ID, analyte, treatment, period — those column names carry through unchanged into `as.data.frame(o_nca)`. They become the basis for `USUBJID`, `PPCAT` (analyte/treatment), and other PP domain identifiers.

```{r grouping-cols}
# Grouping columns appear alongside PPTESTCD and PPORRES
df |>
  select(Subject, start, end, PPTESTCD, PPORRES) |>
  head(6)
```

For multi-analyte or multi-period studies, add those variables to the formula in `PKNCAconc()`:

```{r multi-group-example, eval=FALSE}
# Example: analyte + period grouping
d_conc_multi <- PKNCAconc(
  my_data,
  conc ~ Time | USUBJID / ANALYTE / PERIOD
)
```

All grouping variables will appear as columns in the long-format result, ready to be mapped to their CDISC equivalents.

---

## 3. Extracting Results for Submission

`as.data.frame(o_nca)` returns the individual (subject-level) long-format table. This is the primary output for PP domain construction:

```{r extract-individual}
pp_domain_draft <- as.data.frame(o_nca)

# Rename to CDISC column names as needed
pp_cdisc <- pp_domain_draft |>
  rename(
    PPORRES  = PPORRES,    # already CDISC-named
    PPTESTCD = PPTESTCD    # already CDISC-named
  ) |>
  # PPSTRESN is typically the same as PPORRES for numeric results
  mutate(PPSTRESN = PPORRES)

head(pp_cdisc)
```

The result is in tidy long format: one row per subject per parameter per interval, which is the structure expected by the CDISC PP domain.

---

## 4. The `exclude` Column

The `exclude` column records whether a result was flagged for exclusion and why. When `exclude` is `NA`, the result is valid. When it is a non-`NA` character string, it contains a human-readable exclusion reason.

```{r exclude-col}
# All possible exclusion reasons in this result
table(is.na(df$exclude))

# Show any excluded rows
df |> filter(!is.na(exclude)) |> head()
```

Exclusions are set programmatically via the `exclude_nca_*` family of functions (e.g., `exclude_nca_max.aucinf.pext()`, `exclude_nca_min.hl.r.squared()`), or manually via `exclude()`. For regulatory submissions, include all rows in the PP domain — both included and excluded — and map the `exclude` string to `PPEXCLFL = "Y"` with the reason preserved in a supplemental qualifier.

```{r exclude-nca-example}
# Apply standard exclusion rules and inspect excluded rows
o_nca_excl <- o_nca |>
  exclude(reason = "r.squared < 0.9",   FUN = exclude_nca_min.hl.r.squared(0.9)) |>
  exclude(reason = "%AUCextrap > 20%",  FUN = exclude_nca_max.aucinf.pext(20))

df_excl <- as.data.frame(o_nca_excl)
df_excl |> filter(!is.na(exclude)) |> select(Subject, PPTESTCD, PPORRES, exclude) |> head()
```

---

## 5. Summary Table for Reports

`summary(o_nca)` returns a wide-format table suitable for clinical study reports. Each row is a parameter and each column is a summary statistic (geometric mean, CV%, range, etc.):

```{r summary-table}
summary(o_nca)
```

The summary uses the `PKNCA.set.summary()` configuration for each parameter type. You can customize which statistics are reported and their formatting by calling `PKNCA.set.summary()` before running `pk.nca()`.

---

## 6. Reproducibility: Pinning Options and Environment

For regulatory analyses, the PKNCA option set used during the run must be documented. `PKNCA.options()` with no arguments returns the full current option set:

```{r options-snapshot}
# Capture the full option set at the start of an analysis script
analysis_options <- PKNCA.options()

# Confirm a key option
analysis_options$auc.method      # trapezoidal method
analysis_options$conc.blq        # BLQ handling
```

Set any non-default options at the top of the analysis script, before any data is processed:

```{r set-options, eval=FALSE}
# Example: use log-linear trapezoid instead of linear-up/log-down
PKNCA.options(auc.method = "lin-log")
```

For full reproducibility, record the session environment alongside the option snapshot:

```{r session-info}
# Save at the end of the analysis script
sessionInfo()
```

For validated environments, use `renv` to lock package versions:

```{r renv-example, eval=FALSE}
# Lock the current environment
renv::snapshot()

# Restore a locked environment on a validation system
renv::restore()
```

Together, `PKNCA.options()`, `sessionInfo()`, and `renv` provide the three layers of reproducibility documentation expected for regulatory NCA submissions.


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pk.nca()](https://humanpred.github.io/pknca/reference/pk.nca.html) · [PKNCA.options()](https://humanpred.github.io/pknca/reference/PKNCA.options.html)
:::
