---
title: "Superposition"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)

d_conc <- as.data.frame(Theoph) |> rename(time = Time, subject = Subject)
d_dose <- Theoph |> as.data.frame() |>
  group_by(Subject) |>
  summarise(dose = Dose[1] * Wt[1], .groups = "drop") |>
  rename(subject = Subject) |>
  mutate(time = 0)
```

## What superposition does

**Superposition** predicts steady-state concentration-time profiles from a single-dose profile by linearly summing lagged copies of the single-dose curve. This is the NCA equivalent of multi-dose simulation, and it requires no compartmental model.

The core assumption is **linear, time-invariant pharmacokinetics**: the drug does not accumulate non-linearly, and clearance and volume do not change over time.

```{mermaid}
flowchart LR
    A["Single dose\nprofile (0→∞)"] --> B["Shift by τ,\nshift by 2τ, ..."]
    B --> C["Sum contributions\nat each time t"]
    C --> D["Predicted SS\nprofile (0→τ)"]
```

---

## Basic superposition

```{r}
# Use Theoph subject 1 as single-dose reference
subj1 <- d_conc |> filter(subject == "1")
actual_dose <- d_dose |> filter(subject == "1") |> pull(dose)

o_conc_1 <- PKNCAconc(subj1, conc ~ time | subject)

# Predict steady-state profile with τ = 24 h, show 3 dosing intervals
ss_profile <- superposition(
  o_conc_1,
  tau         = 24,           # dosing interval (h)
  dose.input  = actual_dose,  # dose used to generate the single-dose data
  dose.amount = actual_dose,  # dose to simulate at SS (same here)
  n.tau       = 3,            # simulate 3 intervals
  check.blq   = FALSE         # Theoph has non-zero first sample
)

head(ss_profile, 12)
```

---

## Visualising the approach to steady state

With `n.tau = Inf`, superposition runs until the profile converges to steady state (within `steady.state.tol`).

```{r}
ss_converged <- superposition(
  o_conc_1,
  tau         = 24,
  dose.input  = actual_dose,
  dose.amount = actual_dose,
  n.tau       = Inf,
  check.blq   = FALSE
)

ggplot(ss_converged, aes(x = time, y = conc)) +
  geom_line(colour = "steelblue") +
  geom_point(size = 2, colour = "steelblue") +
  labs(title = "Superposition: predicted steady-state profile (τ = 24 h)",
       x = "Time (h)", y = "Predicted SS concentration (mg/L)") +
  theme_minimal()
```

---

## Changing dose at steady state

To predict the effect of a dose change, set `dose.amount` to the new dose while keeping `dose.input` as the original dose. PKNCA scales the concentrations proportionally.

```{r}
ss_half_dose <- superposition(
  o_conc_1,
  tau         = 24,
  dose.input  = actual_dose,
  dose.amount = actual_dose / 2,   # half dose
  n.tau       = Inf,
  check.blq   = FALSE
)

bind_rows(
  ss_converged   |> mutate(regimen = "Full dose"),
  ss_half_dose   |> mutate(regimen = "Half dose")
) |>
  ggplot(aes(x = time, y = conc, colour = regimen)) +
  geom_line() + geom_point(size = 2) +
  labs(title = "Dose scaling via superposition",
       x = "Time (h)", y = "Predicted SS concentration (mg/L)") +
  scale_colour_manual(values = c("Full dose" = "steelblue", "Half dose" = "coral")) +
  theme_minimal()
```

---

## Changing the dosing interval (τ)

```{r}
ss_8h <- superposition(
  o_conc_1,
  tau         = 8,
  dose.input  = actual_dose,
  dose.amount = actual_dose,
  n.tau       = Inf,
  check.blq   = FALSE
)

ggplot(ss_8h, aes(x = time, y = conc)) +
  geom_line(colour = "firebrick") +
  geom_point(size = 2, colour = "firebrick") +
  labs(title = "Superposition with τ = 8 h (more frequent dosing)",
       x = "Time within interval (h)", y = "Predicted SS concentration (mg/L)") +
  theme_minimal()
```

---

## Extracting accumulation ratio

The **accumulation ratio (Rac)** is Cmax at steady state divided by Cmax after the first dose:

```{r}
cmax_sd <- max(subj1$conc, na.rm = TRUE)
cmax_ss <- max(ss_converged$conc, na.rm = TRUE)
rac      <- cmax_ss / cmax_sd
cat("Accumulation ratio (Rac):", round(rac, 2), "\n")
cat("Single-dose Cmax:", round(cmax_sd, 2), "mg/L\n")
cat("Steady-state Cmax:", round(cmax_ss, 2), "mg/L\n")
```

---

## Key arguments reference

| Argument | Required | Meaning |
|---|---|---|
| `tau` | Yes | Dosing interval (same time units as data) |
| `dose.input` | When `dose.amount` ≠ NULL | Dose used to generate the observed profile |
| `dose.amount` | When `dose.input` ≠ NULL | Dose to simulate (for dose scaling) |
| `n.tau` | No (default `Inf`) | Number of intervals to simulate before stopping |
| `steady.state.tol` | No (default 0.001) | Convergence tolerance for `n.tau = Inf` |
| `check.blq` | No (default `TRUE`) | Require first concentration = 0; set `FALSE` if first sample is non-zero |
| `auc.type` | No (default `"AUCinf"`) | How to extrapolate the single-dose tail: `"AUCinf"` or `"AUClast"` |
| `additional.times` | No | Extra timepoints to include in the output |

> **Limitation:** Superposition assumes linear PK. It should not be used for drugs with non-linear kinetics (e.g., Michaelis-Menten elimination, time-varying clearance, or saturable protein binding).


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [superposition()](https://humanpred.github.io/pknca/reference/superposition.html)
:::
