---
title: "Time to Steady State"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## What is time to steady state?

**Time to steady state (TSS)** is the dosing occasion (or clock time) at which pre-dose trough concentrations are no longer meaningfully rising. PKNCA provides two NCA-based methods for estimating TSS from trough concentration data collected across multiple dose occasions.

Both methods work on trough concentrations (Ctrough = pre-dose concentration) collected at successive dose numbers, not full PK profiles.

---

## Simulated dataset

We simulate trough concentrations for 6 subjects across 10 dose occasions. Concentrations approach an individual steady-state plateau following a monoexponential rise.

```{r}
set.seed(42)
n_subjects <- 6
n_doses    <- 10

d_tss <- data.frame(
  Subject     = rep(1:n_subjects, each = n_doses),
  dose_number = rep(1:n_doses, n_subjects),
  ke          = rep(0.1 * exp(rnorm(n_subjects, 0, 0.2)), each = n_doses),
  Css         = rep(10  * exp(rnorm(n_subjects, 0, 0.3)), each = n_doses)
)
d_tss$ctrough <- d_tss$Css * (1 - exp(-d_tss$ke * d_tss$dose_number * 24)) +
                 rnorm(nrow(d_tss), 0, 0.1)

ggplot(d_tss, aes(x = dose_number, y = ctrough, group = Subject, colour = factor(Subject))) +
  geom_line() + geom_point() +
  labs(title = "Simulated trough concentrations across dose occasions",
       x = "Dose number", y = "Ctrough (mg/L)", colour = "Subject") +
  theme_minimal()
```

---

## Method 1: Monoexponential (`pk.tss.monoexponential`)

Fits a monoexponential approach-to-plateau model:

$$C_{\text{trough}}(n) = C_{ss} \cdot \left(1 - e^{-k_{\text{tss}} \cdot n}\right)$$

Reports TSS as the dose occasion at which the predicted concentration reaches a specified fraction of Css (default: 90% = `tss.fraction = 0.9`).

Returns population, population-individual (popind), individual, and single-subject estimates.

```{r}
tss_mono <- pk.tss.monoexponential(
  conc        = d_tss$ctrough,
  time        = d_tss$dose_number,
  subject     = d_tss$Subject,
  time.dosing = 1:n_doses
)

tss_mono
```

**Interpreting the output columns:**

| Column | Meaning |
|---|---|
| `tss.monoexponential.population` | TSS from population-level fit (one value shared across subjects) |
| `tss.monoexponential.popind` | TSS from mixed model with subject-level random effect for TSS (NA if not supported by data) |
| `tss.monoexponential.individual` | TSS from individual-level fits |
| `tss.monoexponential.single` | TSS ignoring between-subject variability |

### Changing the fraction threshold

```{r}
# 95% of Css
pk.tss.monoexponential(
  conc        = d_tss$ctrough,
  time        = d_tss$dose_number,
  subject     = d_tss$Subject,
  time.dosing = 1:n_doses,
  tss.fraction = 0.95
)
```

---

## Method 2: Stepwise linear (`pk.tss.stepwise.linear`)

A non-parametric approach: iteratively tests whether concentrations are still increasing linearly from dose occasion $k$ onward. TSS is the earliest dose occasion at which the upward trend is no longer statistically significant.

This method is more robust when the rise to steady state is irregular or non-monoexponential.

```{r}
tss_linear <- pk.tss.stepwise.linear(
  conc        = d_tss$ctrough,
  time        = d_tss$dose_number,
  subject     = d_tss$Subject,
  time.dosing = 1:n_doses
)

tss_linear
```

The result is a data frame with a `tss.stepwise.linear` column containing the first dose occasion number where concentrations are deemed flat (or `NA` if steady state cannot be determined from the data).

---

## Combined method (`pk.tss`)

`pk.tss()` runs both methods and returns results side by side:

```{r}
tss_combined <- pk.tss(
  conc        = d_tss$ctrough,
  time        = d_tss$dose_number,
  subject     = d_tss$Subject,
  time.dosing = 1:n_doses
)

tss_combined
```

---

## Key arguments

| Argument | Default | Meaning |
|---|---|---|
| `conc` | — | Trough concentrations (one per dose occasion per subject) |
  | `time` | — | Dose occasion numbers or actual clock times |
| `subject` | — | Subject identifiers |
| `time.dosing` | — | All possible dosing occasions in the study (needed to define the full grid) |
| `tss.fraction` | `0.9` | Fraction of Css that defines "at steady state" (monoexponential method) |
| `output` | `c("population","popind","individual","single")` | Which output types to return |
| `verbose` | `FALSE` | Print model fitting details |

> **Note:** TSS estimation requires **multiple dose occasions with consistent trough sampling**. It is not applicable to single-dose studies or studies where trough collection was inconsistent.


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [pk.tss()](https://humanpred.github.io/pknca/reference/pk.tss.html) · [pk.tss.monoexponential()](https://humanpred.github.io/pknca/reference/pk.tss.monoexponential.html) · [pk.tss.stepwise.linear()](https://humanpred.github.io/pknca/reference/pk.tss.stepwise.linear.html)
:::
