---
title: "Units Assignment and Conversion"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## Why units matter

By default PKNCA returns bare numbers. When you provide units, every row in
`as.data.frame()` gains CDISC-style columns that mirror the SDTM PK-Parameters
domain:

| Column | Meaning |
|--------|---------|
| `PPORRESU` | Original reporting unit (the unit the result is expressed in) |
| `PPSTRESU` | Standardised unit (populated when a conversion is requested) |
| `PPSTRES` | Result expressed in the standardised unit |

This makes the output submission-ready and prevents silent unit mismatches when
combining results from multiple studies or analytes.

---

## Assigning units at object creation

Units can be supplied directly to `PKNCAconc()` and `PKNCAdose()` as string
literals or as the name of a column in your data frame that holds the unit
string per row.

### Arguments in `PKNCAconc()`

Verified with:

```r
formals(getS3method("PKNCAconc", "data.frame")) |> names()
```

The unit-related arguments are:

- `concu` — concentration unit (e.g. `"mg/L"`)
- `timeu` — time unit (e.g. `"h"`)
- `amountu` — amount unit, used for urine/fecal data (e.g. `"mg"`)
- `concu_pref` — preferred output unit for concentration (triggers conversion)
- `timeu_pref` — preferred output unit for time (triggers conversion)
- `amountu_pref` — preferred output unit for amount (triggers conversion)

### Arguments in `PKNCAdose()`

- `doseu` — dose unit (e.g. `"mg"`)
- `doseu_pref` — preferred output unit for dose (triggers conversion)

### Example: units as string literals

```{r}
Theoph_df <- as.data.frame(Theoph)
d_dose <- Theoph_df |>
  dplyr::filter(Time == 0) |>
  dplyr::mutate(amt = Dose * Wt)

o_conc <- PKNCAconc(Theoph_df, conc ~ Time | Subject,
                    concu = "mg/L", timeu = "h")

o_dose <- PKNCAdose(d_dose, amt ~ Time | Subject,
                    doseu = "mg")
```

### Example: units from a data column

```{r}
Theoph_units <- Theoph_df |>
  dplyr::mutate(conc_unit = "mg/L", time_unit = "h")

o_conc_col <- PKNCAconc(Theoph_units, conc ~ Time | Subject,
                        concu = "conc_unit", timeu = "time_unit")
```

---

## `pknca_units_table()`

This function builds a complete mapping from every NCA parameter to its
reporting unit, derived from the four base units you supply.

In PKNCA ≥ 0.12.2, `pknca_units_table()` is an **S3 generic**. The most useful new method is `pknca_units_table.PKNCAdata()`, which extracts units directly from a fully assembled `PKNCAdata` object — useful when units are already embedded in `PKNCAconc` and `PKNCAdose`:

```{r}
# Units already embedded in the data objects
o_conc_u <- PKNCAconc(as.data.frame(Theoph) |> dplyr::rename(time=Time, subject=Subject),
                       conc ~ time | subject, concu = "mg/L", timeu = "h")
o_dose_u <- PKNCAdose(
  as.data.frame(Theoph) |>
    dplyr::group_by(Subject) |>
    dplyr::summarise(dose = Dose[1]*Wt[1], .groups="drop") |>
    dplyr::rename(subject=Subject) |>
    dplyr::mutate(time = 0),
  dose ~ time | subject, route = "extravascular",
  doseu = "mg", timeu = "h"
)
o_data_u <- PKNCAdata(o_conc_u, o_dose_u)

# Extract units from the PKNCAdata object (≥ 0.12.2)
u_from_data <- pknca_units_table(o_data_u)
head(u_from_data, 6)
```

### Arguments (default method)

```{r}
args(getS3method("pknca_units_table", "default"))
```

All four of `concu`, `doseu`, `amountu`, `timeu` are required. PKNCA derives
composite units automatically (e.g. `auclast` gets `h*mg/L` from `timeu="h"`
and `concu="mg/L"`).

### Output columns

The returned data frame has exactly two columns when no preferred units are
requested:

| Column | Content |
|--------|---------|
| `PPORRESU` | Reporting unit for this parameter |
| `PPTESTCD` | Parameter code (e.g. `"cmax"`, `"auclast"`, `"tmax"`) |

When preferred units are requested, two additional columns appear:
`PPSTRESU` (target unit) and `conversion_factor`.

```{r}
u <- pknca_units_table(concu = "mg/L", timeu = "h",
                       doseu = "mg", amountu = "mg")

# The table covers all NCA parameters
nrow(u)

# Sample rows
u[u$PPTESTCD %in% c("cmax", "auclast", "tmax", "cl.obs", "vz.obs"), ]
```

---

## Passing the units table to `PKNCAdata()`

Pass the table via the `units` argument. If `units` is provided, any units
given directly to `PKNCAconc()` or `PKNCAdose()` are ignored.

```{r}
# Build objects (units here will be overridden by the table below)
o_conc2 <- PKNCAconc(Theoph_df, conc ~ Time | Subject)
o_dose2  <- PKNCAdose(d_dose, amt ~ Time | Subject)

u <- pknca_units_table(concu = "mg/L", timeu = "h",
                       doseu = "mg", amountu = "mg")

o_data <- PKNCAdata(o_conc2, o_dose2, units = u)

# Run NCA
res <- pk.nca(o_data)

# Units appear in the PPORRESU column
df <- as.data.frame(res)
df |>
  dplyr::filter(Subject == "1",
                PPTESTCD %in% c("auclast", "cmax", "tmax", "cl.obs")) |>
  dplyr::select(PPTESTCD, PPORRES, PPORRESU)
```

---

## Preferred units

Setting `concu_pref`, `timeu_pref`, `doseu_pref`, or `amountu_pref` in
`pknca_units_table()` tells PKNCA to convert results to those preferred units.
PKNCA calculates the conversion factor automatically for metric prefixes and
time unit changes.

When preferred units are active the results data frame gains:

- `PPSTRESU` — the preferred unit
- `PPSTRES` — the result in the preferred unit

```{r}
u_pref <- pknca_units_table(
  concu = "mg/L", timeu  = "h",
  doseu = "mg",   amountu = "mg",
  concu_pref = "ug/L",   # mg/L → ug/L  (× 1 000)
  timeu_pref = "day"     # h   → day   (× 1/24)
)

# Inspect what the table says for key parameters
u_pref[u_pref$PPTESTCD %in% c("cmax", "auclast", "tmax"), ]
```

```{r}
o_data_pref <- PKNCAdata(o_conc2, o_dose2, units = u_pref)
res_pref <- pk.nca(o_data_pref)

as.data.frame(res_pref) |>
  dplyr::filter(Subject == "1",
                PPTESTCD %in% c("auclast", "cmax", "tmax")) |>
  dplyr::select(PPTESTCD, PPORRES, PPORRESU, PPSTRES, PPSTRESU)
```

`PPORRES`/`PPORRESU` hold the original value and unit; `PPSTRES`/`PPSTRESU`
hold the converted value and unit.

---

## Custom unit conversions

For conversions PKNCA cannot derive automatically (e.g. mass ↔ molar, or a
non-standard reporting preference), pass a `conversions` data frame to
`pknca_units_table()`.

### Required columns

| Column | Type | Meaning |
|--------|------|---------|
| `PPORRESU` | character | The original unit to match |
| `PPSTRESU` | character | The target unit |
| `conversion_factor` | numeric | Multiplier (original × factor = target). Use `NA` to ask PKNCA to compute it. |

### Example: report concentration as ug/mL instead of mg/L

`mg/L` and `ug/mL` are numerically equal (factor = 1), but differ in label.

```{r}
my_conversions <- data.frame(
  PPORRESU        = c("mg/L",   "h*mg/L"),
  PPSTRESU        = c("ug/mL",  "h*ug/mL"),
  conversion_factor = c(1,       1)
)

u_custom <- pknca_units_table(
  concu = "mg/L", timeu = "h",
  doseu = "mg",   amountu = "mg",
  conversions = my_conversions
)

o_data_custom <- PKNCAdata(o_conc2, o_dose2, units = u_custom)
res_custom    <- pk.nca(o_data_custom)

as.data.frame(res_custom) |>
  dplyr::filter(Subject == "1",
                PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::select(PPTESTCD, PPORRES, PPORRESU, PPSTRES, PPSTRESU)
```

---

## Verifying units in results

Call `as.data.frame()` on the `PKNCAresults` object. When a units table was
provided, the `PPORRESU` column is always populated; `PPSTRESU` and `PPSTRES`
appear only when preferred or custom conversions are in effect.

```{r}
df_all <- as.data.frame(res)

# Confirm PPORRESU is present for every row
stopifnot("PPORRESU" %in% names(df_all))

# Browse the first several parameter results for subject 1
df_all |>
  dplyr::filter(Subject == "1") |>
  dplyr::select(PPTESTCD, PPORRES, PPORRESU) |>
  head(10)
```

---

## Summary

| Goal | How |
|------|-----|
| Assign units at data creation | `concu`, `timeu`, `doseu` args in `PKNCAconc()` / `PKNCAdose()` |
| Build a complete unit table | `pknca_units_table(concu, doseu, amountu, timeu)` |
| Pass units to the analysis | `PKNCAdata(..., units = u)` |
| Convert to preferred output units | Add `concu_pref`, `timeu_pref`, etc. to `pknca_units_table()` |
| Custom / non-metric conversions | `conversions` argument in `pknca_units_table()` |
| Inspect units in results | `as.data.frame(res)$PPORRESU` |
| Extract units from a PKNCAdata object (≥ 0.12.2) | `pknca_units_table(o_data)` |
| Allow some parameters without units (≥ 0.12.2) | `PKNCA.options(allow_partial_missing_units = TRUE)` |


---

::: {.callout-note icon=false appearance="minimal"}
**pkgdown reference:** [PKNCAconc()](https://humanpred.github.io/pknca/reference/PKNCAconc.html) · [PKNCAdose()](https://humanpred.github.io/pknca/reference/PKNCAdose.html) · [PKNCAdata()](https://humanpred.github.io/pknca/reference/PKNCAdata.html) · [pknca_units_table()](https://humanpred.github.io/pknca/reference/pknca_units_table.html) · [PKNCA.options()](https://humanpred.github.io/pknca/reference/PKNCA.options.html)
:::
