---
title: "Utility Functions"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## Overview

PKNCA exports several lower-level utility functions useful for data preparation, interval validation, and direct computation outside the full `PKNCAdata` → `pk.nca()` pipeline.

---

## BLQ concentration handling: `clean.conc.blq()`

Applies the BLQ-handling rule to a concentration vector **before** fitting. The full analysis pipeline applies this automatically, but `clean.conc.blq()` is useful for pre-processing or testing your own rule.

```{r}
conc <- c(0.5, 0.3, 0,    0.1, 0,    0.0, 0.02)
time <- c(0,   1,   2,    3,   4,    5,   6   )

# Drop BLQ (0) values
clean.conc.blq(conc, time, conc.blq = "drop")
```

```{r}
# Replace BLQ with half the LLOQ (= 0.05 here)
clean.conc.blq(conc, time, conc.blq = 0.05)
```

The `conc.blq` argument accepts:
- `"drop"` — remove BLQ rows
- `"keep"` — keep BLQ rows as-is
- A number — substitute BLQ values with that number

---

## NA concentration handling: `clean.conc.na()`

Removes or replaces NA concentrations. Internally called before AUC computation.

```{r}
conc_na <- c(0.5, NA, 0.3, 0.1, NA)
time_na <- c(0,   1,   2,   3,   4)

# Default: drop NAs
clean.conc.na(conc_na, time_na, na.treatment = "drop")
```

```{r}
# Replace NAs with 0
clean.conc.na(conc_na, time_na, na.treatment = 0)
```

---

## Concentration interpolation / extrapolation: `interp.extrap.conc()`

Interpolates or extrapolates a concentration at a target time given a known concentration-time profile. Used internally when an interval boundary falls between two observed timepoints.

```{r}
conc_ex <- c(10, 8, 5, 2.5, 1)
time_ex <- c(0,  1, 2, 4,   8)

# Interpolate at t = 3 (between t=2 and t=4)
interp.extrap.conc(conc = conc_ex, time = time_ex, time.out = 3)
```

```{r}
# Extrapolate beyond the last sample (t = 12)
interp.extrap.conc(conc = conc_ex, time = time_ex, time.out = 12)
```

The method (linear vs log-linear) follows the `auc.method` setting. Extrapolation uses the terminal log-linear slope (requires at least 3 points for λz estimation).

---

## Direct half-life computation: `pk.calc.half.life()`

Runs the full curve-stripping algorithm on raw concentration-time data and returns the selected half-life, λz, and associated diagnostics. Useful for batch processing outside the standard pipeline.

```{r}
subj1 <- as.data.frame(Theoph) |> filter(Subject == "1")

hl_result <- pk.calc.half.life(
  conc    = subj1$conc,
  time    = subj1$Time,
  options = PKNCA.options()
)

hl_result[c("half.life", "lambda.z", "r.squared", "lambda.z.n.points",
            "lambda.z.time.first", "span.ratio")]
```

This is the same function called internally by `pk.nca()` when `half.life = TRUE`.

---

## Direct AUC computation: `pk.calc.auc.all()`

Computes AUC (trapezoid rule) directly on a concentration-time vector. Returns the scalar AUC value.

```{r}
pk.calc.auc.all(
  conc   = subj1$conc,
  time   = subj1$Time,
  method = "lin up/log down"
)
```

Other direct AUC functions:

| Function | AUC type |
|---|---|
| `pk.calc.auc.last()` | AUC to last measurable concentration |
| `pk.calc.auc.all()` | AUC to last, treating BLQ as 0 |
| `pk.calc.auc.inf.obs()` | AUC extrapolated to ∞ (observed Clast) |
| `pk.calc.auc.inf.pred()` | AUC extrapolated to ∞ (predicted Clast) |

These are the building blocks behind every AUC parameter in the registry.

---

## Validating interval specifications: `check.interval.specification()`

Validates an intervals data frame before passing it to `PKNCAdata()`. Returns the data frame with all parameter columns filled in (missing columns defaulted to `FALSE`).

```{r}
my_intervals <- data.frame(
  start    = 0,
  end      = Inf,
  auclast  = TRUE,
  cmax     = TRUE
)

# Returns the full expanded interval spec
expanded <- check.interval.specification(my_intervals)

# Which parameters are requested?
names(expanded)[as.logical(expanded[1, ])]
```

```{r}
# Invalid specification — missing start column — raises an error
tryCatch(
  check.interval.specification(data.frame(end = Inf, auclast = TRUE)),
  error = function(e) cat("Error:", conditionMessage(e), "\n")
)
```

This is useful for debugging custom interval definitions before running the full analysis.

---

## Listing all available parameters

```{r}
cols <- get.interval.cols()
cat("Total registered parameters:", length(cols), "\n")

# Show first few with their dependency lists
lapply(head(cols, 5), function(x) x$depends)
```

```{r}
# All dose-normalized parameters
dn_params <- names(cols)[endsWith(names(cols), ".dn")]
cat("Dose-normalized parameters:\n")
cat(paste(" -", dn_params), sep = "\n")
```

---

## Modifying PKNCAdose after creation: `setRoute()` and `setDuration()`

After creating a `PKNCAdose` object, you can change the route or infusion duration without rebuilding from scratch.

```{r}
d_dose_u <- data.frame(subject = "1", dose = 100, time = 0)
o_dose_u <- PKNCAdose(d_dose_u, dose ~ time | subject, route = "extravascular")

# Change to intravascular
o_dose_iv <- setRoute(o_dose_u, route = "intravascular")

# Add infusion duration (0.5 h = 30 min)
o_dose_inf <- setDuration(o_dose_iv, duration = 0.5)

# Verify
o_dose_inf$data[c("route", "duration")]
```

This is useful when you want to test how route assumption changes affect MRT or Vss calculations without rebuilding the full `PKNCAdata` object.

---

## Getting parameter dependencies: `get.parameter.deps()`

Returns the upstream parameters that a given parameter depends on. Useful when debugging why a parameter is `NA` (a dependency may have failed).

```{r}
get.parameter.deps("vss.obs")
```

```{r}
get.parameter.deps("half.life")
```

---

## Listing all current option values: `PKNCA.options()`

`PKNCA.options()` with no arguments returns the full named list of current settings:

```{r}
PKNCA.options()
```

> **Note:** `PKNCA.options.describe()` is exported but non-functional in PKNCA 0.12.1 (raises an error). Use `PKNCA.options()` to inspect current settings, and see the [Workflow Overview](workflow.qmd) page for the full option reference table.

---

## Choosing an option value: `PKNCA.choose.option()`

Selects a value from either the local `options` list passed to `PKNCAdata()` or the global `PKNCA.options()` defaults. Useful when writing custom parameter functions that need to respect session-level settings.

```{r}
# With no local override, returns the global default
PKNCA.choose.option("min.hl.points", options = list())
```

```{r}
# With a local override, the local value wins
PKNCA.choose.option("min.hl.points", options = list(min.hl.points = 5))
```
