---
title: "Validation and Testing"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

PKNCA ships with a comprehensive `testthat`-based test suite that covers the full calculation engine. This page explains how to access and run that suite, how to interpret results, and how to spot-check individual calculations against hand-derived values.

---

## 1. PKNCA's Validation Approach

The validation strategy is **reproducibility-first**: re-running the test suite in your local environment and obtaining the same results as the original package development constitutes confirmation that the software is working correctly in your system.

The test suite uses the [`testthat`](https://testthat.r-lib.org/) framework and targets near-complete coverage of every calculation function. Tests range from numerical spot-checks of individual PK parameters to integration tests of the full `pk.nca()` pipeline.

To check whether tests were installed alongside the package:

```{r test-location}
system.file("tests", package = "PKNCA")
```

If this returns an empty string, the tests were not installed. Standard CRAN installs do not include tests by default; see Section 2 for how to install them.

---

## 2. Installing with Tests and Running the Suite

Tests must be explicitly included at install time.

**From CRAN (source install with tests):**

```{r cran-install, eval=FALSE}
install.packages(
  pkgs         = "PKNCA",
  INSTALL_opts = "--install-tests",
  type         = "source"
)
```

**From GitHub:**

```{r github-install, eval=FALSE}
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github(
  "humanpred/pknca",
  INSTALL_opts = "--install-tests"
)
```

**Running the test suite** once tests are installed:

```{r run-tests, eval=FALSE}
if (!requireNamespace("testthat")) install.packages("testthat")

# Option A — via devtools (recommended during development)
devtools::test(system.file(package = "PKNCA"))

# Option B — via testthat directly
testthat::test_package("PKNCA")

# Option C — full R CMD CHECK (most rigorous; runs examples too)
# From a terminal:
# R CMD CHECK $(R -e 'cat(find.package("PKNCA"))')
```

---

## 3. Interpreting Results

A passing run reports the number of tests, the number of expectations checked, and zero failures. A typical run looks like:

```
[ FAIL 0 | WARN 0 | SKIP n | PASS N ]
```

- **FAIL 0** — no calculation errors; the install is valid.
- **SKIP n** — some tests are conditionally skipped (e.g., tests that require internet access, or environment-specific checks). Skips are expected and do not indicate a problem.
- **WARN** — warnings during tests are expected; the PKNCA validation vignette explicitly notes that warnings appear during testing and are not displayed in the published vignette.

If any failures are reported, the output shows the test file, test name, and the specific expectation that failed — compare the `expected` and `actual` values to diagnose whether the issue is a platform difference or a genuine calculation error.

---

## 4. Checking the Installed Version

Before any regulatory analysis, record the exact package version:

```{r version-check}
packageVersion("PKNCA")
```

Cite the package version in the analysis report:

```{r citation}
citation("PKNCA")
```

The canonical citation is:

> Denney W, Duvvuri S, Buckeridge C (2015). "Simple, Automatic Noncompartmental Analysis: The PKNCA R Package." *Journal of Pharmacokinetics and Pharmacodynamics*, 42(1), 11-107,S65. doi:10.1007/s10928-015-9432-2

---

## 5. Spot-Checking a Calculation Against a Known Reference

Running the full test suite confirms the package as a whole. For a targeted validation of a specific parameter, you can compute the value by hand and compare it to `pk.nca()` output.

**Example: manual linear-trapezoid AUClast for Theoph Subject 1**

The linear (additive) trapezoid rule is:

$$\text{AUC}_{t_i \to t_{i+1}} = \frac{(t_{i+1} - t_i)(C_i + C_{i+1})}{2}$$

```{r spot-check}
data("Theoph")

# Subject 1 concentration-time data
s1 <- Theoph[Theoph$Subject == "1", c("Time", "conc")]
s1

# Manual linear-trapezoid AUClast
# AUClast uses all data up to and including the last measurable concentration
n <- nrow(s1)
manual_auclast <- sum(
  (s1$Time[2:n] - s1$Time[1:(n - 1)]) *
    (s1$conc[2:n] + s1$conc[1:(n - 1)]) / 2
)
cat("Manual AUClast (linear trap):", round(manual_auclast, 3), "\n")
```

Now run `pk.nca()` and extract the same parameter:

```{r pknca-auclast}
d_conc <- PKNCAconc(Theoph, conc ~ Time | Subject)
d_dose <- PKNCAdose(
  Theoph |> group_by(Subject) |>
    summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>
    mutate(Time = 0),
  Dose ~ Time | Subject
)

# Use a single interval that spans the full profile
d_data <- PKNCAdata(
  d_conc, d_dose,
  intervals = data.frame(start = 0, end = Inf, auclast = TRUE)
)
suppressMessages(o_nca <- pk.nca(d_data))

pknca_auclast <- as.data.frame(o_nca) |>
  filter(Subject == "1", PPTESTCD == "auclast") |>
  pull(PPORRES)

cat("PKNCA AUClast:              ", round(pknca_auclast, 3), "\n")
cat("Difference (manual - PKNCA):", round(manual_auclast - pknca_auclast, 6), "\n")
```

The two values should agree to numerical precision. PKNCA uses the linear-up / log-down trapezoid method by default (controlled by `PKNCA.options("auc.method")`); for an all-linear comparison, either set `PKNCA.options(auc.method = "linear")` or use only time points where the log-down rule would not apply (i.e., where concentrations are not monotonically decreasing between consecutive points).

To confirm the current method:

```{r auc-method}
PKNCA.options("auc.method")
```

---

## 6. Reproducibility Reporting with `sessionInfo()`

Every validation run should capture the complete session environment. Record this at the end of each analysis script:

```{r session-info}
Sys.Date()
sessionInfo()
```

`sessionInfo()` captures:

- R version and platform
- Locale and timezone
- All attached and loaded packages with their exact versions
- The version of PKNCA and all its dependencies

For validated regulatory environments, supplement `sessionInfo()` with an `renv` lockfile:

```{r renv-lock, eval=FALSE}
# Capture the environment
renv::snapshot()

# Confirm a clean restore on the validation system
renv::restore()
renv::status()
```

Together, `packageVersion("PKNCA")`, `PKNCA.options()`, `sessionInfo()`, and `renv::snapshot()` provide a complete and reproducible record of the computational environment used in an NCA analysis.
