---
title: "Workflow Overview"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
```

## The Five-Step Workflow

Every PKNCA analysis follows the same five steps regardless of route or complexity.

```{mermaid}
flowchart TD
    S1["① Prepare data\nconcentration data.frame\ndose data.frame"] --> S2
    S2["② PKNCAconc()\ndefine conc column, time column,\ngrouping variables, route-specific args"] --> S3
    S3["③ PKNCAdose()\ndefine dose, time, route\n(intravascular / extravascular)"] --> S4
    S4["④ PKNCAdata()\ncombine conc + dose\nspecify intervals + imputation + units"] --> S5
    S5["⑤ pk.nca()\nrun all calculations"] --> S6
    S6["PKNCAresults\nsummary() / as.data.frame()"]

    style S1 fill:#f0f4ff
    style S6 fill:#e8f5e9
```

## Step-by-step: minimal example

Below is the complete workflow with the built-in `Theoph` dataset (oral theophylline, 12 subjects).

### Step 1 — Prepare your data

PKNCA expects two separate data frames: one for concentrations and one for doses. Both must be in **long format** (one row per observation).

```{r}
# Built-in oral theophylline dataset
head(Theoph)
```

The columns are: `Subject`, `Wt` (weight, kg), `Dose` (mg/kg), `Time` (h), `conc` (mg/L).

```{r}
# Concentration data: keep what we need
d_conc <- Theoph |>
  select(Subject, Time, conc)

# Dose data: one row per subject (single dose at time 0)
d_dose <- Theoph |>
  group_by(Subject) |>
  summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>  # convert mg/kg to mg
  mutate(Time = 0)

head(d_dose)
```

### Step 2 — PKNCAconc

`PKNCAconc` tells PKNCA the structure of your concentration data via a formula:
`concentration ~ time | group1 / group2 / ...`

```{r}
o_conc <- PKNCAconc(d_conc, conc ~ Time | Subject)
print(o_conc)
```

### Step 3 — PKNCAdose

`PKNCAdose` describes the dosing. The `route` argument is critical — it determines which parameters are estimable.

```{r}
o_dose <- PKNCAdose(d_dose, Dose ~ Time | Subject, route = "extravascular")
print(o_dose)
```

### Step 4 — PKNCAdata

`PKNCAdata` combines concentration and dose objects. If you don't specify `intervals`, PKNCA automatically determines them from the dosing times.

```{r}
o_data <- PKNCAdata(o_conc, o_dose)
```

You can inspect the auto-generated intervals:

```{r}
o_data$intervals |> head()
```

### Step 5 — pk.nca

Run all calculations:

```{r}
o_nca <- pk.nca(o_data)
```

### Results

```{r}
# Tidy data frame of all computed parameters
as.data.frame(o_nca) |> head(20)
```

```{r}
# Summary table (geometric mean ± CV% by default for most parameters)
summary(o_nca)
```

---

## Specifying intervals manually

Auto-interval detection works for simple designs. For complex studies (multiple doses, custom windows), define intervals yourself.

An **interval** is a row in a data frame with:
- `start`, `end`: time boundaries
- One column per parameter set to `TRUE` to request that parameter

```{r}
# Request specific parameters over a specific window
my_intervals <- data.frame(
  start  = 0,
  end    = Inf,
  auclast  = TRUE,
  aucinf.obs = TRUE,
  cmax   = TRUE,
  tmax   = TRUE,
  half.life = TRUE
)

o_data2 <- PKNCAdata(o_conc, o_dose, intervals = my_intervals)
o_nca2  <- pk.nca(o_data2)
as.data.frame(o_nca2)
```

---

## Controlling calculations with PKNCA.options()

`PKNCA.options()` is the global settings registry. Call with no arguments to see all current settings.

```{r}
PKNCA.options()
```

Key options and what they control:

| Option | Default | Effect |
|---|---|---|
| `auc.method` | `"lin up/log down"` | AUC integration method |
| `min.hl.points` | `3` | Minimum points required for half-life fit |
| `min.hl.r.squared` | `0.9` | Minimum adjusted R² for half-life acceptance |
| `max.aucinf.pext` | `20` | Maximum % extrapolation allowed for AUCinf |
| `conc.na` | `"drop"` | How to handle NA concentrations |
| `first.tmax` | `TRUE` | Use first vs. last Tmax when tied |
| `allow.tmax.in.half.life` | `FALSE` | Allow Tmax point in λz regression |

Override for a single analysis by passing `options` to `PKNCAdata()`:

```{r}
o_data_strict <- PKNCAdata(
  o_conc, o_dose,
  options = list(
    min.hl.points  = 4,     # require at least 4 points for half-life
    min.hl.r.squared = 0.95 # stricter R² threshold
  )
)
```

---

## Extracting and plotting results

```{r}
results_df <- as.data.frame(o_nca)

# Filter to one parameter
auclast_df <- results_df |>
  filter(PPTESTCD == "auclast")

ggplot(auclast_df, aes(x = Subject, y = PPORRES)) +
  geom_col(fill = "steelblue") +
  labs(title = "AUClast by Subject", x = "Subject", y = "AUClast (h·mg/L)") +
  theme_minimal()
```

---

## Updating results

Use `update()` to recalculate with modified settings without rebuilding from scratch:

```{r}
# Change AUC method and recalculate
o_nca_updated <- update(o_nca, data = update(o_data, options = list(auc.method = "linear")))
```
