---
title: "Workflow Overview"
---

```{r setup, include=FALSE}
library(PKNCA)
library(dplyr)
library(ggplot2)
conflicted::conflicts_prefer(dplyr::filter, dplyr::select, .quiet = TRUE)
```

## The Five-Step Workflow

Every PKNCA analysis follows the same five steps regardless of route or complexity.

```{mermaid}
flowchart TD
    S1["① Prepare data\nconcentration data.frame\ndose data.frame"] --> S2
    S2["② PKNCAconc()\ndefine conc column, time column,\ngrouping variables, route-specific args"] --> S3
    S3["③ PKNCAdose()\ndefine dose, time, route\n(intravascular / extravascular)"] --> S4
    S4["④ PKNCAdata()\ncombine conc + dose\nspecify intervals + imputation + units"] --> S5
    S5["⑤ pk.nca()\nrun all calculations"] --> S6
    S6["PKNCAresults\nsummary() / as.data.frame()"]

    style S1 fill:#f0f4ff
    style S6 fill:#e8f5e9
```

## Step-by-step: minimal example

Below is the complete workflow with the built-in `Theoph` dataset (oral theophylline, 12 subjects).

### Step 1 — Prepare your data

PKNCA expects two separate data frames: one for concentrations and one for doses. Both must be in **long format** (one row per observation).

```{r}
# Built-in oral theophylline dataset
head(Theoph)
```

The columns are: `Subject`, `Wt` (weight, kg), `Dose` (mg/kg), `Time` (h), `conc` (mg/L).

```{r}
# Concentration data: keep what we need
d_conc <- Theoph |>
  select(Subject, Time, conc)

# Dose data: one row per subject (single dose at time 0)
d_dose <- Theoph |>
  group_by(Subject) |>
  summarise(Dose = Dose[1] * Wt[1], .groups = "drop") |>  # convert mg/kg to mg
  mutate(Time = 0)

head(d_dose)
```

### Step 2 — PKNCAconc

`PKNCAconc` tells PKNCA the structure of your concentration data via a formula:
`concentration ~ time | group1 / group2 / ...`

```{r}
o_conc <- PKNCAconc(d_conc, conc ~ Time | Subject)
print(o_conc)
```

### Step 3 — PKNCAdose

`PKNCAdose` describes the dosing. The `route` argument is critical — it determines which parameters are estimable.

```{r}
o_dose <- PKNCAdose(d_dose, Dose ~ Time | Subject, route = "extravascular")
print(o_dose)
```

### Step 4 — PKNCAdata

`PKNCAdata` combines concentration and dose objects. If you don't specify `intervals`, PKNCA automatically determines them from the dosing times.

```{r}
o_data <- PKNCAdata(o_conc, o_dose)
```

You can inspect the auto-generated intervals:

```{r}
o_data$intervals |> head()
```

### Step 5 — pk.nca

Run all calculations:

```{r}
o_nca <- pk.nca(o_data)
```

### Results

```{r}
# Tidy data frame of all computed parameters
as.data.frame(o_nca) |> head(20)
```

```{r}
# Summary table (geometric mean ± CV% by default for most parameters)
summary(o_nca)
```

---

## Specifying intervals manually

Auto-interval detection works for simple designs. For complex studies (multiple doses, custom windows), define intervals yourself.

An **interval** is a row in a data frame with:
- `start`, `end`: time boundaries
- One column per parameter set to `TRUE` to request that parameter

```{r}
# Request specific parameters over a specific window
my_intervals <- data.frame(
  start  = 0,
  end    = Inf,
  auclast  = TRUE,
  aucinf.obs = TRUE,
  cmax   = TRUE,
  tmax   = TRUE,
  half.life = TRUE
)

o_data2 <- PKNCAdata(o_conc, o_dose, intervals = my_intervals)
o_nca2  <- pk.nca(o_data2)
as.data.frame(o_nca2)
```

---

## Controlling calculations with PKNCA.options()

`PKNCA.options()` is the global settings registry. Call with no arguments to see all current settings.

```{r}
PKNCA.options()
```

All options and what they control:

**AUC / integration**

| Option | Default | Effect |
|---|---|---|
| `auc.method` | `"lin up/log down"` | Integration rule: `"linear"`, `"lin-log"`, or `"lin up/log down"` |

**Half-life / λz**

| Option | Default | Effect |
|---|---|---|
| `min.hl.points` | `3` | Minimum data points for a valid λz regression |
| `min.hl.r.squared` | `0.9` | Minimum adjusted R² for λz acceptance |
| `adj.r.squared.factor` | `0.0001` | Selection tolerance: the longest regression window whose adj.R² is within this value of the best adj.R² is chosen (favours more points when fits are nearly equal) |
| `min.span.ratio` | `2` | Minimum ratio of time span to half-life; rejects regressions over too short a window |
| `allow.tmax.in.half.life` | `FALSE` | Allow Tmax timepoint in λz regression |
| `hl_method` | `"log-linear"` | `"log-linear"` or `"tobit"` — **not in CRAN 0.12.1, requires GitHub version** |
| `tobit_n_points_penalty` | `0` | Extra penalty per point in tobit — **GitHub version only** |

**AUCinf extrapolation**

| Option | Default | Effect |
|---|---|---|
| `max.aucinf.pext` | `20` | Flag results where extrapolated % of AUCinf exceeds this threshold |

**Concentration handling**

| Option | Default | Effect |
|---|---|---|
| `conc.na` | `"drop"` | NA concentrations: `"drop"` removes them; a number substitutes that value |
| `conc.blq` | `list(first="keep", middle="drop", last="keep")` | BLQ concentrations: `"drop"`, `"keep"`, or a number globally; or a named list with keys `first` (before first non-BLQ), `middle` (between non-BLQs), `last` (after last non-BLQ), or `before.tmax`/`after.tmax` |

**Tmax / Tmin**

| Option | Default | Effect |
|---|---|---|
| `first.tmax` | `TRUE` | `TRUE` = return first Tmax when tied; `FALSE` = return last |

**Summary / output**

| Option | Default | Effect |
|---|---|---|
| `max.missing` | `0.5` | Maximum fraction of missing values allowed before suppressing a summary statistic |
| `keep_interval_cols` | `NULL` | Additional columns from the intervals data frame to carry through to results |
| `allow_partial_missing_units` | `FALSE` | Allow some parameters to lack unit definitions when a units table is provided |
| `single.dose.aucs` | (list) | Default interval specifications used when no intervals are provided for a single dose |
| `tau.choices` | `NA` | Candidate dosing intervals (τ) for auto-detection in multiple-dose designs |
| `progress` | `TRUE` | Show progress bar during `pk.nca()` |
| `debug` | `NULL` | Enable internal debug output (not for routine use) |

Override for a single analysis by passing `options` to `PKNCAdata()`:

```{r}
o_data_strict <- PKNCAdata(
  o_conc, o_dose,
  options = list(
    min.hl.points  = 4,     # require at least 4 points for half-life
    min.hl.r.squared = 0.95 # stricter R² threshold
  )
)
```

---

## Extracting and plotting results

```{r}
results_df <- as.data.frame(o_nca)

# Filter to one parameter
auclast_df <- results_df |>
  filter(PPTESTCD == "auclast")

ggplot(auclast_df, aes(x = Subject, y = PPORRES)) +
  geom_col(fill = "steelblue") +
  labs(title = "AUClast by Subject", x = "Subject", y = "AUClast (h·mg/L)") +
  theme_minimal()
```

---

## Re-running with different options

PKNCA does not have an `update()` method. To rerun with changed settings,
pass a new `options` list to `PKNCAdata()` and call `pk.nca()` again:

```{r}
# Change AUC method and recalculate
o_data_linear <- PKNCAdata(o_conc, o_dose, options = list(auc.method = "linear"))
o_nca_linear  <- pk.nca(o_data_linear)
```
